\documentclass[11pt,a4paper]{article}

% -----------------------------------------------------------------------------
% Packages
% -----------------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc, fit, matrix, decorations.pathreplacing, shapes.geometric, backgrounds}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[most]{tcolorbox}
\usepackage{caption}
\usepackage{float}

\geometry{margin=2.5cm}
\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=green!45!black,
  urlcolor=blue!60!black
}

% -----------------------------------------------------------------------------
% Theorem environments
% -----------------------------------------------------------------------------
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\newtheorem{lemma}{Lemma}[section]

% -----------------------------------------------------------------------------
% Boxes
% -----------------------------------------------------------------------------
\newtcolorbox{motivationbox}[1][]{
  colback=green!3!white,
  colframe=green!45!black,
  fonttitle=\bfseries,
  title={#1},
  breakable
}

\newtcolorbox{intuitionbox}[1][]{
  colback=purple!3!white,
  colframe=purple!45!black,
  fonttitle=\bfseries,
  title={#1},
  breakable
}

\newtcolorbox{connectionbox}[1][]{
  colback=cyan!3!white,
  colframe=cyan!55!black,
  fonttitle=\bfseries,
  title={#1},
  breakable
}

\newtcolorbox{tradeoffbox}[1][]{
  colback=red!2!white,
  colframe=red!60!black,
  fonttitle=\bfseries,
  title={#1},
  breakable
}

\newtcolorbox{warningbox}[1][]{
  colback=orange!4!white,
  colframe=orange!70!black,
  fonttitle=\bfseries,
  title={#1},
  breakable
}

% -----------------------------------------------------------------------------
% Code listing style
% -----------------------------------------------------------------------------
\definecolor{codebg}{RGB}{245,245,245}
\definecolor{codeframe}{RGB}{200,200,200}
\lstset{
    language=Python,
    basicstyle=\ttfamily\small,
    backgroundcolor=\color{codebg},
    frame=single,
    rulecolor=\color{codeframe},
    keywordstyle=\color{blue!70!black}\bfseries,
    stringstyle=\color{red!60!black},
    commentstyle=\color{green!50!black}\itshape,
    showstringspaces=false,
    breaklines=true,
    numbers=left,
    numberstyle=\tiny\color{gray},
    xleftmargin=2em,
    framexleftmargin=1.5em,
}

% -----------------------------------------------------------------------------
% Custom commands
% -----------------------------------------------------------------------------
\newcommand{\vx}{\bm{x}}
\newcommand{\vu}{\bm{u}}
\newcommand{\vg}{\bm{g}}
\newcommand{\vq}{\bm{q}}
\newcommand{\vk}{\bm{k}}
\newcommand{\vv}{\bm{v}}
\newcommand{\vo}{\bm{o}}
\newcommand{\vw}{\bm{w}}
\newcommand{\vh}{\bm{h}}
\newcommand{\vy}{\bm{y}}
\newcommand{\vz}{\bm{z}}
\newcommand{\vtheta}{\bm{\theta}}
\newcommand{\mW}{\bm{W}}
\newcommand{\mQ}{\bm{Q}}
\newcommand{\mK}{\bm{K}}
\newcommand{\mV}{\bm{V}}
\newcommand{\mA}{\bm{A}}
\newcommand{\mR}{\bm{R}}
\newcommand{\mS}{\bm{S}}
\newcommand{\mC}{\bm{C}}
\newcommand{\mI}{\bm{I}}
\newcommand{\mM}{\bm{M}}
\newcommand{\mX}{\bm{X}}
\newcommand{\mY}{\bm{Y}}
\newcommand{\mH}{\bm{H}}
\newcommand{\mO}{\bm{O}}
\newcommand{\mP}{\bm{P}}
\newcommand{\mZ}{\bm{Z}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\dmodel}{d_{\mathrm{model}}}
\newcommand{\dhead}{d_{\mathrm{head}}}
\newcommand{\dff}{d_{\mathrm{ff}}}
\newcommand{\nhead}{n_{\mathrm{head}}}
\newcommand{\nlayer}{n_{\mathrm{layer}}}
\newcommand{\vocab}{|\mathcal{V}|}
\newcommand{\loss}{\mathcal{L}}
\newcommand{\data}{\mathcal{D}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\given}{\,|\,}
\DeclareMathOperator{\softmax}{softmax}
\DeclareMathOperator{\ReLU}{ReLU}
\DeclareMathOperator{\GELU}{GELU}
\DeclareMathOperator{\SiLU}{SiLU}
\DeclareMathOperator{\ELU}{ELU}
\DeclareMathOperator{\RMS}{RMS}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\KL}{KL}
\DeclareMathOperator{\CE}{CE}
\DeclareMathOperator{\argmax}{arg\,max}

% -----------------------------------------------------------------------------
% TikZ styles
% -----------------------------------------------------------------------------
\tikzset{
  block/.style={draw, rounded corners, thick, align=center, minimum height=8mm, minimum width=22mm, fill=blue!6},
  data/.style={draw, rounded corners, thick, align=center, minimum height=7mm, minimum width=18mm, fill=green!7},
  tensor/.style={draw, rounded corners, thick, align=center, fill=purple!7, minimum height=7mm},
  op/.style={draw, circle, thick, inner sep=1.2pt, fill=white},
  smallblock/.style={draw, rounded corners, thick, align=center, font=\small, fill=blue!5, minimum height=6mm},
  pressure/.style={draw, rounded corners, thick, align=center, fill=orange!8, minimum height=7mm},
  replacement/.style={draw, rounded corners, thick, align=center, fill=green!8, minimum height=7mm},
  arrow/.style={-{Latex[length=2mm]}, thick},
  residual/.style={-{Latex[length=2.3mm]}, very thick, blue!70!black},
  grad/.style={-{Latex[length=2.3mm]}, very thick, red!70!black, dashed},
  note/.style={align=center, font=\small},
  every node/.style={font=\small}
}

% -----------------------------------------------------------------------------
% Title
% -----------------------------------------------------------------------------
\title{\textbf{Transformer Variant Lab: Detailed Learning Notes}\\[0.5em]
\large Mathematical, Statistical, and Systems Concepts Across Decoder-Only Variants V0--V6}
\author{Project Notes -- Mathematical Statistics and Machine Learning Level}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

% =============================================================================
\section{Introduction and Organization}
% =============================================================================

This document explains the mathematical, statistical, and computational concepts used across the decoder-only Transformer variants implemented in the project. It deliberately avoids code-level and API-level detail. Each concept has one home section containing its definition, motivation, mechanism, worked intuition, equations, and trade-offs:
\begin{itemize}
    \item \textbf{V0 (Vanilla):} GPT-2-style decoder with learned absolute positions, LayerNorm, multi-head attention, and a ReLU/GELU FFN.
    \item \textbf{V1 (Modern):} RoPE, RMSNorm, SwiGLU, and memory-efficient exact attention.
    \item \textbf{V2 (ALiBi):} replaces rotary position geometry with head-specific linear distance biases.
    \item \textbf{V3 (GQA):} reduces the number of key/value heads while retaining many query heads.
    \item \textbf{V4 (SWA):} restricts selected attention layers to a local causal window; an interleaved sub-variant alternates local and global layers.
    \item \textbf{V5 (Linear attention):} replaces the softmax similarity kernel with a positive feature map that admits a recurrent prefix summary.
    \item \textbf{V6 (MoE):} replaces dense FFNs with sparsely routed expert FFNs so total capacity can grow faster than per-token computation.
\end{itemize}

The goal is to explain the \emph{reasoning chain} behind the architectural shifts, not merely to list them. The notes therefore use the same pattern inside each component family:
\begin{equation*}
    \text{baseline role}
    \rightarrow
    \text{limitation}
    \rightarrow
    \text{design pressure}
    \rightarrow
    \text{replacement mechanism}
    \rightarrow
    \text{new trade-off}.
\end{equation*}

\begin{motivationbox}[One topic, one home section]
A previous draft repeated the same explanations in a component-shift section, a technical section, and a final roadmap. This version removes that redundancy. Each major topic is discussed in exactly one home section:
\begin{itemize}[nosep]
    \item residual stream and block equations: \cref{sec:block-residual};
    \item statistical objective: \cref{sec:statistical-objective};
    \item normalization and residual stability: \cref{sec:normalization};
    \item attention, GQA, RoPE, ALiBi, FlashAttention, sliding windows, linear attention, and attention entropy: \cref{sec:attention-position};
    \item ReLU/GELU/SiLU, SwiGLU, and sparse mixture-of-experts layers: \cref{sec:ffn};
    \item KV-cache and decoding: \cref{sec:generation};
    \item bias, dropout, and weight tying: \cref{sec:simplifications};
    \item optimization, precision, fair comparison, hardware efficiency, and representation diagnostics: \cref{sec:training-eval}.
\end{itemize}
Later sections may refer back to these topics, but they do not re-explain them.
\end{motivationbox}

\subsection{Notation}

Throughout, we use the following conventions:
\begin{itemize}[nosep]
    \item $B$ is batch size, $T$ is sequence length, and $\dmodel$ is the residual-stream width.
    \item $\nhead$ is the number of query heads, $\dhead=\dmodel/\nhead$ is the dimension per head, and $n_{\mathrm{kv}}$ is the number of key/value heads in GQA.
    \item $\dff$ is the standard FFN hidden dimension, and $d_h$ is the SwiGLU or expert hidden dimension.
    \item $\nlayer$ is the number of Transformer blocks, $W$ is a sliding-window length, and $r$ denotes a linear-attention feature dimension when it differs from $\dhead$.
    \item $N_{\mathrm{exp}}$ is the number of MoE experts and $k$ is the number selected for one token.
    \item $\mathcal{V}$ is the vocabulary, and $\vocab$ is its size.
    \item Bold lowercase symbols such as $\vx,\vq,\vk$ denote vectors. Bold uppercase symbols such as $\mW,\mQ,\mK$ denote matrices or tensors.
    \item $\odot$ denotes element-wise multiplication, and $\sigma(x)=(1+e^{-x})^{-1}$ is the logistic sigmoid.
\end{itemize}

\subsection{Tensor-shape conventions}

For a mini-batch of token IDs $\bm{s}\in\{1,\ldots,\vocab\}^{B\times T}$, the embedding layer returns
\begin{equation}
    \mX^{(0)}\in\reals^{B\times T\times \dmodel}.
\end{equation}
For attention, query, key, and value tensors are usually reshaped to
\begin{equation}
    \mQ,\mK,\mV\in\reals^{B\times\nhead\times T\times\dhead}.
\end{equation}
The final logits have shape
\begin{equation}
    \mZ\in\reals^{B\times T\times\vocab}.
\end{equation}
Keeping shapes explicit prevents many implementation mistakes, especially when adding RoPE or a KV-cache.

% =============================================================================
\section{Statistical Objective}
\label{sec:statistical-objective}
% =============================================================================

\subsection{Autoregressive factorization}

A decoder-only language model defines a probability distribution over token sequences $s_{1:T}$ by the chain rule:
\begin{equation}\label{eq:chain_rule_lm}
    p_{\vtheta}(s_{1:T})=
    \prod_{t=1}^{T}p_{\vtheta}(s_t\given s_{<t}),
\end{equation}
where $s_{<t}=(s_1,\ldots,s_{t-1})$ and $\vtheta$ denotes all trainable parameters. Causal masking is the architectural enforcement of this conditioning restriction: the prediction at position $t$ must not use tokens $s_{t+1},\ldots,s_T$.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=9mm and 8mm]
    \node[data] (tokens) {token IDs\\$s_{1:T}$};
    \node[block, right=of tokens] (emb) {embedding};
    \node[block, right=of emb] (blocks) {$\nlayer$ decoder\\blocks};
    \node[block, right=of blocks] (head) {LM head\\logits};
    \node[data, right=of head] (prob) {softmax\\$p_{\vtheta}(s_{t+1}\mid s_{\le t})$};
    \node[data, below=9mm of prob] (target) {shifted targets\\$s_{2:T+1}$};
    \node[block, left=of target] (loss) {cross-entropy\\loss};

    \draw[arrow] (tokens) -- (emb);
    \draw[arrow] (emb) -- (blocks);
    \draw[arrow] (blocks) -- (head);
    \draw[arrow] (head) -- (prob);
    \draw[arrow] (prob) -- (loss);
    \draw[arrow] (target) -- (loss);
    \draw[grad] (loss.south) .. controls +(0,-1.0) and +(0,-1.0) .. node[below, note] {backpropagation updates all parameters} (emb.south);
\end{tikzpicture}
\caption{Training flow for a decoder-only language model. V0 and V1 change the internal parameterization of the decoder blocks, but the objective remains shifted-token maximum likelihood.}
\label{fig:training-flow}
\end{figure}

Given a dataset $\data=\{s^{(i)}_{1:T_i}\}_{i=1}^N$, maximum likelihood estimation minimizes the negative log-likelihood:
\begin{equation}\label{eq:nll}
    \loss(\vtheta)=
    -\sum_{i=1}^{N}\sum_{t=1}^{T_i}
    \log p_{\vtheta}\!\left(s^{(i)}_t\given s^{(i)}_{<t}\right).
\end{equation}
The average token-level loss is
\begin{equation}\label{eq:avg_nll}
    \bar{\loss}(\vtheta)=
    -\frac{1}{M}\sum_{i=1}^{N}\sum_{t=1}^{T_i}
    \log p_{\vtheta}\!\left(s^{(i)}_t\given s^{(i)}_{<t}\right),
    \qquad M=\sum_{i=1}^{N}T_i.
\end{equation}

\subsection{Cross-entropy and KL divergence}

Let $q$ be the true conditional next-token distribution and $p_{\vtheta}$ the model distribution. For a fixed context,
\begin{equation}\label{eq:cross_entropy_def}
    H(q,p_{\vtheta})
    =-\sum_{w\in\mathcal{V}}q(w)\log p_{\vtheta}(w).
\end{equation}
The KL decomposition is
\begin{equation}\label{eq:entropy_decomposition}
    H(q,p_{\vtheta})=H(q)+\KL(q\Vert p_{\vtheta}).
\end{equation}
Since $H(q)$ does not depend on $\vtheta$, minimizing cross-entropy is equivalent, at the population level, to minimizing $\KL(q\Vert p_{\vtheta})$.

In data, we observe one true next token per context. The empirical target is a one-hot distribution
\begin{equation}
    \hat q(w)=\ind[w=s^{\mathrm{true}}_t],
\end{equation}
so the cross-entropy collapses to
\begin{equation}
    H(\hat q,p_{\vtheta})=-\log p_{\vtheta}(s^{\mathrm{true}}_t\given s_{<t}).
\end{equation}
This is the per-token training loss used in code.

\begin{motivationbox}[Why the statistical objective matters]
The architecture is not directly trained to ``sound good.'' It is trained to assign high probability to the next token under held-out contexts. A component is useful if it lowers validation cross-entropy, reaches the same loss faster, or computes the same probability distribution with less memory and time.
\end{motivationbox}

\subsection{Perplexity}

Perplexity is the exponential of the average NLL:
\begin{equation}\label{eq:perplexity}
    \mathrm{PPL}=\exp(\bar{\loss}).
\end{equation}
If the average loss is measured in nats, perplexity can be interpreted as an effective branching factor. A perplexity of $20$ means that the model is, on average, about as uncertain as choosing uniformly among $20$ plausible next tokens. This interpretation is approximate because real next-token distributions are not uniform.

\subsection{Empirical risk, validation loss, and tokenization}

Training loss is an empirical estimate of population risk. A separate validation set estimates out-of-sample risk:
\begin{equation}
    \widehat R_{\mathrm{val}}(\vtheta)=
    -\frac{1}{M_{\mathrm{val}}}\sum_{(i,t)\in\mathrm{val}}
    \log p_{\vtheta}(s^{(i)}_t\given s^{(i)}_{<t}).
\end{equation}
The generalization gap is
\begin{equation}
    \widehat R_{\mathrm{val}}(\vtheta)-\widehat R_{\mathrm{train}}(\vtheta).
\end{equation}
A small model can underfit, giving high training and validation loss. A large model trained too long on too little data can overfit, giving low training loss but worse validation loss.

The model distribution is over tokens, not directly over words or characters. Therefore, perplexities are only comparable when the tokenizer, vocabulary, data split, and context length are held fixed.

\begin{warningbox}[Data leakage]
Validation loss is meaningful only if validation text is not duplicated in the training set. If validation documents leak into training, the validation loss becomes an optimistic estimate of generalization.
\end{warningbox}

% =============================================================================
\section{Decoder Block and Residual Stream}
\label{sec:block-residual}
% =============================================================================

\subsection{Input representation}

In V0, the initial residual stream is the sum of token and learned position embeddings:
\begin{equation}\label{eq:v0_embedding}
    \mX^{(0)}_{b,t,:}=\mW_{\mathrm{tok}}[s_{b,t}]+\mW_{\mathrm{pos}}[t].
\end{equation}
In V1, RoPE is applied inside attention, so the initial residual stream uses token embeddings only:
\begin{equation}\label{eq:v1_embedding}
    \mX^{(0)}_{b,t,:}=\mW_{\mathrm{tok}}[s_{b,t}].
\end{equation}

\subsection{Residual stream}

\begin{definition}[Residual stream]
The \emph{residual stream} is the hidden representation of all tokens as it flows through the Transformer:
\begin{equation}
    \mX^{(0)},\mX^{(1)},\ldots,\mX^{(\nlayer)}
    \in\reals^{B\times T\times\dmodel}.
\end{equation}
Each block updates this stream additively:
\begin{equation}
    \mX\leftarrow\mX+\mathrm{Sublayer}(\mathrm{Norm}(\mX)).
\end{equation}
Thus, the residual stream is not a separate parameter matrix. It is the evolving token representation that attention and FFN sublayers repeatedly read from and write to.
\end{definition}

Attention writes cross-token information into the residual stream. The FFN writes nonlinear token-wise features into the same stream. This is why architecture choices interact: if the FFN update becomes stronger, normalization and initialization must still keep the additive stream stable.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.72, transform shape, node distance=7mm and 8mm]
    \node[tensor] (x) {$\mX^{(\ell)}$};
    \node[block, right=of x] (n1) {Norm};
    \node[block, right=of n1] (attn) {causal\\attention};
    \node[op, right=of attn] (p1) {$+$};
    \node[tensor, right=of p1] (y) {$\mY^{(\ell)}$};
    \node[block, right=of y] (n2) {Norm};
    \node[block, right=of n2] (ffn) {FFN};
    \node[op, right=of ffn] (p2) {$+$};
    \node[tensor, right=of p2] (out) {$\mX^{(\ell+1)}$};

    \draw[arrow] (x) -- (n1);
    \draw[arrow] (n1) -- (attn);
    \draw[arrow] (attn) -- (p1);
    \draw[arrow] (p1) -- (y);
    \draw[arrow] (y) -- (n2);
    \draw[arrow] (n2) -- (ffn);
    \draw[arrow] (ffn) -- (p2);
    \draw[arrow] (p2) -- (out);
    \draw[residual] (x.north) .. controls +(0,1.2) and +(0,1.2) .. (p1.north);
    \draw[residual] (y.north) .. controls +(0,1.2) and +(0,1.2) .. (p2.north);
\end{tikzpicture}
\caption{Pre-normalized residual block. Normalization controls the input scale of each sublayer; residual arrows preserve a direct information and gradient path.}
\label{fig:preln-block}
\end{figure}

\subsection{Block equations}

For layer $\ell=0,\ldots,\nlayer-1$:
\begin{align}
    \mY^{(\ell)}
    &=\mX^{(\ell)}+
      \mathrm{Attn}^{(\ell)}\!\left(\mathrm{Norm}^{(\ell)}_1(\mX^{(\ell)})\right),\label{eq:block_attn}\\
    \mX^{(\ell+1)}
    &=\mY^{(\ell)}+
      \mathrm{FFN}^{(\ell)}\!\left(\mathrm{Norm}^{(\ell)}_2(\mY^{(\ell)})\right).\label{eq:block_ffn}
\end{align}
A final normalization is often applied before the output head:
\begin{equation}\label{eq:final_norm_logits}
    \mH=\mathrm{Norm}_{\mathrm{final}}(\mX^{(\nlayer)}),
    \qquad
    \mZ=\mH\mW_{\mathrm{head}}^{\top}.
\end{equation}
The conditional probability is
\begin{equation}\label{eq:lm_softmax}
    p_{\vtheta}(s_{t+1}=w\given s_{\le t})=
    \frac{\exp(Z_{t,w})}{\sum_{u\in\mathcal{V}}\exp(Z_{t,u})}.
\end{equation}

\subsection{Stable softmax}

For numerical stability, softmax is computed after subtracting the row maximum:
\begin{equation}\label{eq:stable_softmax}
    \softmax(\vz)_i=
    \frac{\exp(z_i-c)}{\sum_j\exp(z_j-c)},
    \qquad c=\max_j z_j.
\end{equation}
This does not change the mathematical value because softmax is shift-invariant:
\begin{equation}
    \softmax(\vz+c\mathbf{1})=\softmax(\vz).
\end{equation}
It prevents overflow when logits are large.

\subsection{Parameter-count accounting}

A rough parameter count is useful when comparing variants.
\begin{itemize}[nosep]
    \item \textbf{Embeddings:} token embeddings contain $\vocab\dmodel$ parameters. Learned absolute position embeddings add $T_{\max}\dmodel$ parameters in V0. RoPE adds no trainable parameters in V1.
    \item \textbf{Attention:} standard dense Q, K, V, and output projections contain approximately $4\dmodel^2$ parameters per layer, ignoring biases.
    \item \textbf{Standard FFN:} a two-matrix FFN with $\dff=4\dmodel$ contains $2\dmodel\dff=8\dmodel^2$ parameters per layer.
    \item \textbf{SwiGLU FFN:} a gated FFN with hidden size $d_h$ contains $3\dmodel d_h$ parameters. Parameter matching with the standard FFN gives $d_h\approx(8/3)\dmodel$.
\end{itemize}
Ignoring normalization scales, a decoder layer therefore has about $4\dmodel^2+8\dmodel^2=12\dmodel^2$ non-embedding parameters when the FFN is parameter-matched.

% =============================================================================
\section{Residual Stability and Normalization}
\label{sec:normalization}
% =============================================================================

\subsection{The scaling problem created by residual additions}

A residual block adds a learned update to its input:
\begin{equation}\label{eq:residual}
    \vx^{(\ell+1)}=\vx^{(\ell)}+f^{(\ell)}(\vx^{(\ell)}).
\end{equation}
If residual updates were independent and had coordinate variance $\sigma_f^2$, then after $L$ additions,
\begin{equation}
    \Var[x_i^{(L)}]\approx \Var[x_i^{(0)}]+L\sigma_f^2.
\end{equation}
The independence assumption is only a heuristic, but it shows the risk: residual streams can drift in scale with depth. Normalization and initialization address this problem from opposite sides. Normalization controls the input scale to each sublayer; residual-output initialization controls the update scale written back into the stream.

\subsection{Pre-normalization versus post-normalization}

The original Transformer used post-normalization:
\begin{equation}
    \vx^{(\ell+1)}=\mathrm{Norm}\bigl(\vx^{(\ell)}+f^{(\ell)}(\vx^{(\ell)})\bigr).
\end{equation}
Modern decoder-only language models usually use pre-normalization:
\begin{equation}\label{eq:preln}
    \vx^{(\ell+1)}=\vx^{(\ell)}+f^{(\ell)}(\mathrm{Norm}(\vx^{(\ell)})).
\end{equation}

\begin{proposition}[Direct gradient path in Pre-LN]
For \cref{eq:preln}, the layer Jacobian is
\begin{equation}
    \frac{\partial\vx^{(\ell+1)}}{\partial\vx^{(\ell)}}
    =\mI+J_{f^{(\ell)}}J_{\mathrm{Norm}}.
\end{equation}
The identity term gives a direct residual path for forward activations and backward gradients.
\end{proposition}

\begin{proof}
Differentiate $\vx^{(\ell)}+f^{(\ell)}(\mathrm{Norm}(\vx^{(\ell)}))$ with respect to $\vx^{(\ell)}$ and apply the chain rule.
\end{proof}

This does not guarantee perfect gradient behavior, because products of matrices $\mI+J_fJ_{\mathrm{Norm}}$ can still become poorly conditioned. It does, however, make deep optimization easier than putting normalization directly on the residual path.

\subsection{Component shift: LayerNorm to RMSNorm}

\subsubsection{Baseline role: LayerNorm}

\begin{definition}[Layer Normalization]
For $\vx\in\reals^d$,
\begin{equation}\label{eq:layernorm}
    \mathrm{LayerNorm}(\vx)
    =\bm{\gamma}\odot\frac{\vx-\mu\mathbf{1}}{\sqrt{\sigma^2+\epsilon}}+\bm{\beta},
\end{equation}
where
\begin{equation}
    \mu=\frac{1}{d}\sum_{i=1}^{d}x_i,
    \qquad
    \sigma^2=\frac{1}{d}\sum_{i=1}^{d}(x_i-\mu)^2.
\end{equation}
\end{definition}

LayerNorm performs two operations: it subtracts the feature mean and divides by the feature standard deviation. In V0, this gives each attention and FFN sublayer inputs with predictable coordinate scale.

\subsubsection{Limitation and design pressure}

LayerNorm couples two ideas: re-centering and re-scaling. Re-scaling is directly tied to residual-stream stability: if the input norm doubles, attention logits and FFN activations can become much larger. Re-centering is less obviously essential in pre-normalized decoder-only models, because later linear maps and residual additions can already represent shifted directions. It also adds another reduction over the hidden dimension.

The design pressure is therefore: keep the scale-control part of normalization, simplify the computation, and preserve the pre-norm residual highway.

\subsubsection{Replacement mechanism: RMSNorm}

\begin{definition}[Root Mean Square Normalization]
For $\vx\in\reals^d$,
\begin{equation}\label{eq:rmsnorm}
    \mathrm{RMSNorm}(\vx)=
    \bm{\gamma}\odot\frac{\vx}{\RMS(\vx)},
    \qquad
    \RMS(\vx)=\sqrt{\frac{1}{d}\sum_{i=1}^{d}x_i^2+\epsilon}.
\end{equation}
\end{definition}

RMSNorm removes mean-centering and preserves scale normalization. Since
\begin{equation}\label{eq:rms_variance_relation}
    \RMS(\vx)^2=\frac{1}{d}\Vert\vx\Vert_2^2=\sigma^2+\mu^2,
\end{equation}
RMSNorm is close to LayerNorm when the coordinate mean is small. Unlike LayerNorm, it is not invariant to shifts $\vx\mapsto\vx+c\mathbf{1}$.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.82, transform shape, node distance=6mm and 8mm]
    \node[tensor] (x1) {$\vx$};
    \node[block, right=of x1] (mean) {compute\\mean $\mu$};
    \node[block, right=of mean] (center) {subtract\\$\mu$};
    \node[block, right=of center] (std) {compute\\std $\sigma$};
    \node[block, right=of std] (div1) {divide\\by $\sigma$};
    \node[tensor, right=of div1] (ln) {LayerNorm\\output};

    \node[tensor, below=13mm of x1] (x2) {$\vx$};
    \node[block, right=of x2] (ms) {compute\\mean square};
    \node[block, right=of ms] (rms) {sqrt\\RMS};
    \node[block, right=of rms] (div2) {divide\\by RMS};
    \node[tensor, right=26mm of div2] (rn) {RMSNorm\\output};

    \draw[arrow] (x1) -- (mean);
    \draw[arrow] (mean) -- (center);
    \draw[arrow] (center) -- (std);
    \draw[arrow] (std) -- (div1);
    \draw[arrow] (div1) -- (ln);

    \draw[arrow] (x2) -- (ms);
    \draw[arrow] (ms) -- (rms);
    \draw[arrow] (rms) -- (div2);
    \draw[arrow] (div2) -- (rn);

    \node[note, left=4mm of x1] {V0};
    \node[note, left=4mm of x2] {V1};
    \node[note, below=3mm of rms] {one fewer reduction and no centering step};
\end{tikzpicture}
\caption{LayerNorm versus RMSNorm. RMSNorm keeps scale control but drops explicit mean-centering.}
\label{fig:ln-vs-rms}
\end{figure}

\subsubsection{New trade-off and connections}

RMSNorm is not simply a better LayerNorm. It gives up exact shift invariance. If a large common offset across coordinates becomes harmful, RMSNorm will not remove it. The reason it works well in this design is that pre-normalization mostly needs stable magnitudes before sublayers read from the residual stream.

\begin{connectionbox}[Where the connection belongs]
The main connection is local to this section: RMSNorm controls the input scale to attention and FFN sublayers. Later sections mention this only as a cross-reference. For example, SwiGLU is scale-sensitive because it multiplies two projections, but the detailed explanation of RMSNorm remains here.
\end{connectionbox}

\subsection{RMSNorm Jacobian}

Let
\begin{equation}
    r(\vx)=\sqrt{\frac{1}{d}\Vert\vx\Vert_2^2+\epsilon},
    \qquad
    \vy=\frac{\vx}{r(\vx)}.
\end{equation}
Ignoring the learned scale $\bm{\gamma}$, the Jacobian is
\begin{equation}\label{eq:rms_jacobian}
    \frac{\partial\vy}{\partial\vx}
    =\frac{1}{r}\mI-\frac{1}{d r^3}\vx\vx^\top.
\end{equation}
Thus RMSNorm is not merely division by a constant; it couples coordinates through a rank-one correction.

\subsection{Weight initialization and residual-output initialization}

Initialization is part of the architecture story because the model is optimized from its initial function. A deep residual network can be expressive in principle but unstable in practice if every layer initially writes updates that are too large.

For a linear layer $\vy=\mW\vx$ with independent weights and inputs,
\begin{equation}
    \Var(y_i)=d_{\mathrm{in}}\Var(W_{ij})\Var(x_j).
\end{equation}
This is the motivation behind Xavier/Glorot-style initialization: choose weight variance on the order of $1/d_{\mathrm{in}}$ so that activation variance does not systematically grow or shrink after one matrix multiplication.

\begin{intuitionbox}[What initialization controls]
If weights are too small, all features start near zero and gradients can become weak. If weights are too large, activations and attention logits can saturate. Good initialization makes the first forward pass look like a stable random feature model rather than a numerical accident.
\end{intuitionbox}

GPT-2-style models often initialize most weights as
\begin{equation}
    W_{ij}\sim\mathcal{N}(0,0.02^2),
\end{equation}
with a special depth scaling for residual output projections:
\begin{equation}\label{eq:gpt2_init}
    W_{ij}\sim\mathcal{N}\!\left(0,\frac{0.02^2}{2\nlayer}\right)
\end{equation}
for attention output projections and FFN output projections. The factor $2\nlayer$ appears because each block has two residual additions: one from attention and one from the FFN.

\begin{motivationbox}[Why residual projections are scaled more strongly]
The output projection of attention and the down projection of the FFN are the matrices that write into the residual stream. If each of the $2\nlayer$ writes contributes variance of similar size, the total residual variance can grow with depth. Scaling these projections by roughly $1/\sqrt{2\nlayer}$ makes the initial network closer to an identity map plus many small updates.
\end{motivationbox}

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.90, transform shape, node distance=8mm and 10mm]
    \node[tensor] (x0) {$\vx^{(0)}$};
    \node[op, right=of x0] (p1) {$+$};
    \node[op, right=of p1] (p2) {$+$};
    \node[right=5mm of p2] (dots) {$\cdots$};
    \node[op, right=5mm of dots] (pL) {$+$};
    \node[tensor, right=of pL] (xL) {$\vx^{(L)}$};
    \node[block, below=7mm of p1] (u1) {small residual\\update};
    \node[block, below=7mm of p2] (u2) {small residual\\update};
    \node[block, below=7mm of pL] (uL) {small residual\\update};

    \draw[arrow] (x0) -- (p1);
    \draw[arrow] (p1) -- (p2);
    \draw[arrow] (p2) -- (dots);
    \draw[arrow] (dots) -- (pL);
    \draw[arrow] (pL) -- (xL);
    \draw[arrow] (u1) -- (p1);
    \draw[arrow] (u2) -- (p2);
    \draw[arrow] (uL) -- (pL);
    \node[note, below=6mm of u2] {output projections are scaled by roughly $1/\sqrt{2\nlayer}$ so many additions do not explode at initialization};
\end{tikzpicture}
\caption{Depth-scaled residual projection initialization. The network starts near an identity-plus-small-updates system.}
\label{fig:init-scaling}
\end{figure}

% =============================================================================
\section{Attention, Position, and Attention Computation}
\label{sec:attention-position}
% =============================================================================

\subsection{Self-attention without position is permutation-equivariant}

Without positional information, self-attention is not a sequence model in the usual sense. It is permutation-equivariant: if the input token representations are permuted, the output vectors are permuted in the same way.

\begin{proposition}[Permutation equivariance]
Let $P$ be a permutation matrix applied to the sequence dimension. If no positional information or causal mask is used, then self-attention satisfies
\begin{equation}
    \mathrm{Attn}(P\mX)=P\,\mathrm{Attn}(\mX).
\end{equation}
\end{proposition}
This is why a decoder-only Transformer needs both position information and causal masking.

\subsection{Q, K, and V as content-addressable memory}

\begin{definition}[Scaled dot-product attention]
Given $\mQ\in\reals^{T_q\times\dhead}$, $\mK\in\reals^{T_k\times\dhead}$, and $\mV\in\reals^{T_k\times\dhead}$,
\begin{equation}\label{eq:attention}
    \mathrm{Attention}(\mQ,\mK,\mV)=
    \softmax\!\left(\frac{\mQ\mK^\top}{\sqrt{\dhead}}\right)\mV.
\end{equation}
\end{definition}

Queries and keys decide \emph{where} to read; values contain \emph{what} is read. This separation allows routing and content to use different learned representations.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.68, transform shape, node distance=7mm and 7mm]
    \node[data] (x) {token states\\$\mX$};
    \node[block, right=of x] (wq) {$\mW_Q$};
    \node[tensor, right=of wq] (q) {queries\\$\mQ$};
    \node[block, below=6mm of wq] (wk) {$\mW_K$};
    \node[tensor, right=of wk] (k) {keys\\$\mK$};
    \node[block, below=6mm of wk] (wv) {$\mW_V$};
    \node[tensor, right=of wv] (v) {values\\$\mV$};

    \node[block, right=20mm of k] (score) {match scores\\$\mQ\mK^\top/\sqrt{\dhead}$};
    \node[block, right=of score] (sm) {softmax\\weights};
    \node[block, right=of sm] (read) {weighted read\\$\mP\mV$};
    \node[tensor, right=of read] (out) {attention\\output};

    \draw[arrow] (x) -- (wq);
    \draw[arrow] (x) |- (wk);
    \draw[arrow] (x) |- (wv);
    \draw[arrow] (wq) -- (q);
    \draw[arrow] (wk) -- (k);
    \draw[arrow] (wv) -- (v);
    \draw[arrow] (q) -- (score);
    \draw[arrow] (k) -- (score);
    \draw[arrow] (score) -- (sm);
    \draw[arrow] (sm) -- (read);
    \draw[arrow] (v) -| (read);
    \draw[arrow] (read) -- (out);
\end{tikzpicture}
\caption{Attention as content-addressable memory. Q and K solve the routing problem; V stores the content that is aggregated.}
\label{fig:qkv-memory}
\end{figure}

\subsection{Why divide by \texorpdfstring{$\sqrt{\dhead}$}{sqrt(d-head)}}

Assume $q_i$ and $k_i$ are independent, mean-zero, variance-one random variables. Then
\begin{equation}
    \E[\vq^\top\vk]=0,
    \qquad
    \Var(\vq^\top\vk)=\sum_{i=1}^{\dhead}\Var(q_i k_i)=\dhead.
\end{equation}
Thus the unscaled dot product typically has magnitude $O(\sqrt{\dhead})$. Dividing by $\sqrt{\dhead}$ keeps logits at approximately unit scale and prevents the softmax from saturating at initialization.

\subsection{Causal masking}

For autoregressive language modeling, token $i$ may only attend to tokens $j\leq i$:
\begin{equation}\label{eq:causal_mask}
    A_{ij}=\begin{cases}
        \vq_i^\top\vk_j/\sqrt{\dhead}, & j\leq i,\\
        -\infty, & j>i.
    \end{cases}
\end{equation}
After softmax, entries with $j>i$ receive probability zero.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.95]
    \fill[green!15] (0,0) -- (0,4) -- (4,0) -- cycle;
    \fill[red!10] (0,4) -- (4,4) -- (4,0) -- cycle;
    \draw[step=0.8cm, gray!70] (0,0) grid (4,4);
    \draw[thick] (0,0) rectangle (4,4);
    \node at (1.0,1.0) {allowed};
    \node at (3.0,3.0) {masked};
    \node[below] at (2,-0.25) {key position $j$};
    \node[rotate=90] at (-0.45,2) {query position $i$};
    \node[note] at (2,4.45) {causal mask: each row can only see past and self};
\end{tikzpicture}
\caption{Lower-triangular causal mask. Green entries are visible; red entries are forbidden future context.}
\label{fig:causal-mask}
\end{figure}

\subsection{Multi-head attention}

Each head has its own projections:
\begin{equation}
    \mQ^{(h)}=\mX\mW_Q^{(h)},
    \qquad
    \mK^{(h)}=\mX\mW_K^{(h)},
    \qquad
    \mV^{(h)}=\mX\mW_V^{(h)}.
\end{equation}
Each head computes attention separately, and the outputs are concatenated and projected:
\begin{equation}\label{eq:mha}
    \mathrm{MHA}(\mX)=
    \mathrm{Concat}(\mathrm{head}_1,\ldots,\mathrm{head}_{\nhead})\mW_O.
\end{equation}
Multiple heads let the model learn different retrieval patterns in parallel. The heads are not an ensemble in the statistical sense because they are jointly trained and mixed by $\mW_O$.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.82, transform shape, node distance=7mm and 8mm]
    \node[tensor] (x) {$\mX$};
    \node[block, right=of x] (proj) {QKV\\projection};
    \node[block, right=of proj] (split) {split into\\$\nhead$ heads};
    \node[smallblock, above right=4mm and 10mm of split] (h1) {head 1};
    \node[smallblock, right=10mm of split] (h2) {head 2};
    \node[smallblock, below right=4mm and 10mm of split] (h3) {head $\nhead$};
    \node[block, right=28mm of split] (concat) {concat\\heads};
    \node[block, right=of concat] (wo) {$\mW_O$};
    \node[tensor, right=of wo] (out) {mixed\\output};

    \draw[arrow] (x) -- (proj);
    \draw[arrow] (proj) -- (split);
    \draw[arrow] (split) -- (h1);
    \draw[arrow] (split) -- (h2);
    \draw[arrow] (split) -- (h3);
    \draw[arrow] (h1) -- (concat);
    \draw[arrow] (h2) -- (concat);
    \draw[arrow] (h3) -- (concat);
    \draw[arrow] (concat) -- (wo);
    \draw[arrow] (wo) -- (out);
\end{tikzpicture}
\caption{Multi-head attention. Heads learn different retrieval patterns and the output projection writes the mixed result back into the residual stream.}
\label{fig:multihead}
\end{figure}

\subsection{Grouped-query attention: many queries, fewer key/value memories}
\label{sec:gqa}

\subsubsection{Baseline role and design pressure}

In standard multi-head attention (MHA), each of the $\nhead$ heads has its own query, key, and value representation. This gives every head maximum freedom, but it also stores $\nhead$ key vectors and $\nhead$ value vectors for every cached token. During long-context generation, those K/V tensors can dominate memory traffic and limit the number of sequences that fit on the GPU.

The design pressure is asymmetric: query heads benefit from diversity because they represent different questions asked about the context, while key and value representations can often be shared by several query heads. Grouped-query attention (GQA) keeps $\nhead$ query heads but uses only $n_{\mathrm{kv}}$ key/value heads, where
\begin{equation}
    1 \leq n_{\mathrm{kv}} \leq \nhead,
    \qquad
    g=\frac{\nhead}{n_{\mathrm{kv}}}
\end{equation}
when the heads divide evenly. Here $g$ is the number of query heads served by one K/V head \cite{ainslie2023gqa}.

\subsubsection{Mechanism and tensor shapes}

The projections have shapes
\begin{align}
    \mQ &\in \reals^{B\times\nhead\times T\times\dhead},\\
    \mK,\mV &\in \reals^{B\times n_{\mathrm{kv}}\times T\times\dhead}.
\end{align}
For query head $h\in\{0,\ldots,\nhead-1\}$, define its K/V group
\begin{equation}
    c(h)=\left\lfloor\frac{h}{g}\right\rfloor.
\end{equation}
Then
\begin{equation}\label{eq:gqa-head}
    \mathrm{head}_h
    =\softmax\!\left(
      \frac{\mQ^{(h)}(\mK^{(c(h))})^\top}{\sqrt{\dhead}}
    \right)\mV^{(c(h))}.
\end{equation}
The symbols have the following roles: $h$ identifies a query head, $c(h)$ identifies the shared K/V head used by that query, and $g$ is the number of query heads in each group.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=5mm and 8mm]
    \node[smallblock] (q1) {query 1};
    \node[smallblock, below=of q1] (q2) {query 2};
    \node[smallblock, below=of q2] (q3) {query 3};
    \node[smallblock, below=of q3] (q4) {query 4};
    \node[block, right=20mm of q2, yshift=-3mm, fill=green!8] (kv1) {shared K/V\\head 1};
    \node[smallblock, right=35mm of q1] (q5) {query 5};
    \node[smallblock, below=of q5] (q6) {query 6};
    \node[smallblock, below=of q6] (q7) {query 7};
    \node[smallblock, below=of q7] (q8) {query 8};
    \node[block, right=20mm of q6, yshift=-3mm, fill=green!8] (kv2) {shared K/V\\head 2};
    \foreach \q in {q1,q2,q3,q4}{\draw[arrow] (\q.east) -- (kv1.west);}
    \foreach \q in {q5,q6,q7,q8}{\draw[arrow] (\q.east) -- (kv2.west);}
    \node[note, below=8mm of kv1, xshift=23mm, text width=10cm] {Eight query heads retain eight different retrieval questions, while only two K/V memories are stored. This is GQA with group size $g=4$.};
\end{tikzpicture}
\caption{Grouped-query attention. Several query heads share one key/value head, reducing K/V parameters and cache memory while preserving query diversity.}
\label{fig:gqa-groups}
\end{figure}

\begin{intuitionbox}[Library analogy]
Imagine several researchers using the same library catalogue. Each researcher asks a different question, so the queries remain separate. The catalogue and books, however, need not be duplicated for every researcher. Sharing K/V heads is analogous to sharing the catalogue and collection while retaining different search questions.
\end{intuitionbox}

\subsubsection{Parameter and cache savings}

Ignoring biases, the MHA Q/K/V projections contain
\begin{equation}
    3\dmodel^2
\end{equation}
parameters because $\nhead\dhead=\dmodel$. GQA contains
\begin{equation}\label{eq:gqa-qkv-params}
    \underbrace{\dmodel^2}_{Q}
    +\underbrace{2\dmodel(n_{\mathrm{kv}}\dhead)}_{K,V}
    =\dmodel^2\left(1+2\frac{n_{\mathrm{kv}}}{\nhead}\right).
\end{equation}
The output projection still contributes $\dmodel^2$. The model-wide percentage reduction depends on how much of the model is attention rather than embeddings or FFNs, so it should be computed rather than assumed.

The K/V cache becomes
\begin{equation}\label{eq:gqa-cache}
    2\nlayer B n_{\mathrm{kv}}T\dhead s
\end{equation}
bytes, where $s$ is bytes per scalar. Relative to MHA, the cache reduction factor is
\begin{equation}
    \frac{\nhead}{n_{\mathrm{kv}}}=g.
\end{equation}
For example, $\nhead=8$ and $n_{\mathrm{kv}}=2$ reduce K/V cache memory by a factor of four.

\subsubsection{Spectrum and trade-off}

\begin{center}
\begin{tabular}{lcl}
\toprule
Mechanism & Number of K/V heads & Interpretation\\
\midrule
MHA & $n_{\mathrm{kv}}=\nhead$ & no sharing\\
GQA & $1<n_{\mathrm{kv}}<\nhead$ & sharing within groups\\
MQA & $n_{\mathrm{kv}}=1$ & all queries share one K/V head\\
\bottomrule
\end{tabular}
\end{center}

Reducing $n_{\mathrm{kv}}$ improves cache efficiency but constrains representation: query heads in one group cannot learn independent K/V coordinate systems. GQA is therefore an interpolation between MHA quality and MQA serving efficiency, not a free optimization. RoPE remains valid: all query heads are rotated, and each shared key head is rotated once at its position before being used by the associated query group.


\subsection{Component shift: learned absolute positions to RoPE}

\subsubsection{Baseline role: learned absolute positions}

V0 injects position by adding a learned lookup-table vector to the token embedding:
\begin{equation}
    \vx_t=\mW_{\mathrm{tok}}[s_t]+\mW_{\mathrm{pos}}[t].
\end{equation}
This is simple and effective, but it treats position as an absolute index. The model learns a vector for ``position 57'' rather than a mechanism for ``three tokens before the current token.'' It is also tied to the maximum number of rows in the learned table.

\subsubsection{Design pressure}

Many language patterns are relative: nearby modifiers, matching brackets, indentation, repeated phrases, and earlier references usually depend on offsets or local structure, not absolute sequence indices. Longer contexts add a second pressure: the position mechanism should extend deterministically beyond a fixed embedding table, even if extrapolation quality is not guaranteed.

\subsubsection{Replacement mechanism: RoPE}

RoPE moves positional information from the input embedding sum into the Q/K dot-product geometry \cite{su2021roformer}. For one two-dimensional coordinate pair,
\begin{equation}\label{eq:rope_rotation}
    R(m\theta)=
    \begin{pmatrix}
        \cos(m\theta)&-\sin(m\theta)\\
        \sin(m\theta)&\cos(m\theta)
    \end{pmatrix}.
\end{equation}
For a head of dimension $\dhead$, define frequencies
\begin{equation}\label{eq:rope_freqs}
    \theta_k=\Theta^{-2k/\dhead},
    \qquad k=0,\ldots,\dhead/2-1.
\end{equation}
The full rotation is block diagonal:
\begin{equation}
    \mR_m=\diag(R(m\theta_0),R(m\theta_1),\ldots,R(m\theta_{\dhead/2-1})).
\end{equation}
Queries and keys are rotated as
\begin{equation}
    \widetilde{\vq}_m=\mR_m\vq_m,
    \qquad
    \widetilde{\vk}_n=\mR_n\vk_n.
\end{equation}

\begin{proposition}[Relative offsets in RoPE attention]
For RoPE-rotated queries and keys,
\begin{equation}\label{eq:rope_relative}
    \widetilde{\vq}_m^\top\widetilde{\vk}_n
    =\vq_m^\top \mR_m^\top\mR_n\vk_n
    =\vq_m^\top\mR_{n-m}\vk_n.
\end{equation}
\end{proposition}

\begin{proof}
In each two-dimensional block,
\begin{equation}
    R(m\theta)^\top R(n\theta)=R(-m\theta)R(n\theta)=R((n-m)\theta).
\end{equation}
Applying this to every block gives \cref{eq:rope_relative}.
\end{proof}

\begin{warningbox}[Precise interpretation]
RoPE does not imply that attention scores depend only on distance. The vectors $\vq_m$ and $\vk_n$ still depend on token content and previous layers. RoPE ensures that the positional part of the dot product enters through the relative offset $n-m$.
\end{warningbox}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=7mm and 8mm]
    \node[note] (v0title) {V0: absolute position is added to the residual stream};
    \node[data, below=of v0title] (tok) {token embedding};
    \node[data, below=of tok] (pos) {learned position table row $t$};
    \node[op, right=13mm of $(tok)!0.5!(pos)$] (add) {$+$};
    \node[tensor, right=of add] (x) {input state $\vx_t$};
    \draw[arrow] (tok) -- (add);
    \draw[arrow] (pos) -- (add);
    \draw[arrow] (add) -- (x);

    \node[note, below=18mm of pos] (v1title) {V1: position changes the Q/K attention geometry};
    \node[data, below=of v1title] (tok2) {token embedding};
    \node[block, right=of tok2] (proj) {project to Q and K};
    \node[block, right=of proj] (rot) {rotate by $\mR_t$};
    \node[tensor, right=of rot] (qk) {$\widetilde{\vq}_t,\widetilde{\vk}_t$};
    \draw[arrow] (tok2) -- (proj);
    \draw[arrow] (proj) -- (rot);
    \draw[arrow] (rot) -- (qk);
\end{tikzpicture}
\caption{Absolute embeddings versus RoPE. V0 injects position by addition to the residual stream; V1 injects position into the Q/K dot-product geometry.}
\label{fig:absolute-vs-rope}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=1.15]
    \draw[->, thick] (-0.2,0) -- (2.2,0) node[right] {$x_0$};
    \draw[->, thick] (0,-0.2) -- (0,1.8) node[above] {$x_1$};
    \draw[arrow, blue!70!black] (0,0) -- (1.6,0.35) node[below right] {$\vq$};
    \draw[arrow, red!70!black] (0,0) -- (0.85,1.35) node[above right] {$\mR_m\vq$};
    \draw[->, thick] (0.65,0.14) arc[start angle=12,end angle=58,radius=0.67];
    \node[note] at (0.95,0.48) {$m\theta$};
    \node[note, align=left] at (4.0,0.95) {Each coordinate pair rotates\\by a position-dependent angle.\\Dot products compare phases,\\so relative offsets appear.};
\end{tikzpicture}
\caption{Two-dimensional RoPE rotation for one coordinate pair. The full head vector is partitioned into many such pairs with different frequencies.}
\label{fig:rope-rotation}
\end{figure}

\subsubsection{Frequency choice and long-context behavior}

The frequencies in \cref{eq:rope_freqs} form a geometric progression. Low-index coordinate pairs rotate quickly and can represent fine local offsets. High-index pairs rotate slowly and preserve phase information over longer distances. This multi-frequency design is analogous to using several sinusoidal scales.

RoPE gives a deterministic formula for positions beyond the training context, unlike a finite learned table. However, extrapolation is not guaranteed. High-frequency dimensions wrap many times at long contexts, and the model may see phase patterns outside the training distribution. Practical long-context variants often modify RoPE frequencies or scale positions.

\subsubsection{Why RoPE is applied to Q and K but not V}

Queries and keys determine routing: which token attends to which other token. Values contain content to be aggregated. Applying RoPE to Q and K makes attention scores position-aware. Applying it to V would rotate the content being averaged and is not necessary for relative-position routing.

\subsubsection{Trade-offs and implementation constraints}

RoPE introduces no learned position parameters and gives relative-offset structure in attention scores. The costs are implementation precision and extrapolation risk. The head dimension must be compatible with coordinate pairing, the same pairing convention must be used consistently, and cached keys must be rotated using exactly the same position convention as queries.

\begin{connectionbox}[Connection to no Q/K bias]
A fixed Q/K bias added before RoPE would be rotated by position, creating a position-dependent offset unrelated to token content. This is one reason RoPE-style models often remove Q/K biases. The detailed bias discussion appears once in \cref{sec:simplifications}.
\end{connectionbox}



\subsection{Alternative position mechanism: ALiBi}
\label{sec:alibi}

\subsubsection{Motivation: encode a distance prior directly in the logits}

RoPE modifies query and key geometry. Attention with Linear Biases (ALiBi) takes a simpler route: Q and K remain content projections, and position enters as an additive distance penalty in the attention logits \cite{press2022alibi}. This cleanly separates two terms:
\begin{equation}
    \underbrace{\frac{\vq_i^\top\vk_j}{\sqrt{\dhead}}}_{\text{content compatibility}}
    \quad+\quad
    \underbrace{b_{ij}^{(h)}}_{\text{distance prior}}.
\end{equation}

For a causal head $h$ and an allowed key position $j\leq i$,
\begin{equation}\label{eq:alibi-logit}
    A_{ij}^{(h)}
    =\frac{(\vq_i^{(h)})^\top\vk_j^{(h)}}{\sqrt{\dhead}}
    -m_h(i-j),
\end{equation}
where $i-j\geq0$ is the backward distance and $m_h>0$ is a fixed head-specific slope. Future positions $j>i$ are still assigned $-\infty$ by the causal mask.

\begin{motivationbox}[Why subtract distance?]
Language often has a recency bias: nearby words are more likely to resolve local syntax, agreement, and short references. ALiBi supplies this bias before learning. Content similarity can still overcome it: a distant token with a sufficiently large Q/K match can receive high attention, but it must compensate for the distance penalty.
\end{motivationbox}

\subsubsection{Head-specific horizons}

Different slopes produce different attention horizons. A large $m_h$ heavily penalizes distance and creates a local head. A small $m_h$ creates a broader head. For powers-of-two head counts, the original construction can be written
\begin{equation}\label{eq:alibi-slopes}
    m_h=2^{-8(h+1)/\nhead},
    \qquad h=0,\ldots,\nhead-1,
\end{equation}
up to the chosen ordering of heads. Non-powers-of-two use an interleaving construction rather than applying \cref{eq:alibi-slopes} blindly.

The penalty changes relative odds in a particularly transparent way. For two keys $j_1$ and $j_2$ with equal content logits,
\begin{equation}
    \frac{\alpha_{ij_1}^{(h)}}{\alpha_{ij_2}^{(h)}}
    =\exp\!\left[-m_h\bigl((i-j_1)-(i-j_2)\bigr)\right].
\end{equation}
Thus every extra token of distance multiplies the odds by $e^{-m_h}$ when content evidence is equal.

\begin{example}[Numerical attention horizon]
Suppose two past tokens have identical content scores and one is ten positions farther away. For $m_h=0.5$, the farther token's unnormalized weight is multiplied by $e^{-5}\approx0.0067$ relative to the nearer token. For $m_h=0.02$, the factor is $e^{-0.2}\approx0.82$. The first head is strongly local; the second remains receptive to distant evidence.
\end{example}

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=0.75cm,y=0.62cm]
    \draw[->, thick] (0,0) -- (10.7,0) node[right] {distance $d=i-j$};
    \draw[->, thick] (0,-5.4) -- (0,0.7) node[above] {bias $-m_hd$};
    \draw[very thick, blue!70!black] plot[domain=0:10] (\x,{-0.5*\x});
    \draw[very thick, green!55!black] plot[domain=0:10] (\x,{-0.2*\x});
    \draw[very thick, orange!80!black] plot[domain=0:10] (\x,{-0.05*\x});
    \node[note, blue!70!black] at (7.5,-4.4) {steep: local};
    \node[note, green!55!black] at (8.0,-2.2) {medium};
    \node[note, orange!80!black] at (8.1,-0.75) {shallow: broad};
\end{tikzpicture}
\caption{ALiBi distance penalties. Different heads receive different fixed slopes, producing multiple recency scales without learned position parameters.}
\label{fig:alibi-slopes}
\end{figure}

\subsubsection{Comparison with RoPE and extrapolation}

\begin{center}
\begin{tabular}{p{3.2cm}p{5.0cm}p{5.0cm}}
\toprule
Property & RoPE & ALiBi\\
\midrule
Where position enters & rotations of Q/K coordinates & additive attention-logit bias\\
Relative structure & relative phase in dot products & monotone penalty in relative distance\\
Expressive prior & multiple periodic frequencies & multiple linear recency slopes\\
Cached keys & stored in position-rotated coordinates & stored as raw content projections\\
Beyond training length & formula is defined, but phase distribution shifts & bias is defined, but quality still requires measurement\\
\bottomrule
\end{tabular}
\end{center}

ALiBi was introduced specifically to improve length extrapolation relative to several position baselines. Nevertheless, ``defined for arbitrary distance'' is weaker than ``guaranteed to generalize.'' Longer sequences may still create unseen content patterns and probability calibration problems.

\begin{warningbox}[Corrections to common misconceptions]
RoPE rotations are mathematically invertible because $\mR_t^{-1}=\mR_t^\top$; cached rotated keys are not irreversibly destroyed. The practical distinction is that ALiBi does not require rotating or later unrotating them. Conversely, ALiBi is not more expressive merely because it is simpler: its position prior is monotone and linear, whereas RoPE provides a richer multi-frequency geometry.
\end{warningbox}


\subsection{Component shift: explicit attention matrix to FlashAttention-style kernels}

\subsubsection{Baseline computation}

V0 attention materializes the full attention score matrix.
\begin{algorithm}[H]
\caption{Explicit causal self-attention}
\begin{algorithmic}[1]
\State $[\mQ;\mK;\mV]\gets \mX\mW_{\mathrm{qkv}}$ \Comment{shape $(B,T,3\dmodel)$}
\State reshape each tensor to $(B,\nhead,T,\dhead)$
\State $\mA\gets \mQ\mK^\top/\sqrt{\dhead}$ \Comment{shape $(B,\nhead,T,T)$}
\State set $A_{ij}\gets-\infty$ for $j>i$
\State $\mP\gets\softmax(\mA,\mathrm{dim}=-1)$
\State $\mO\gets\mP\mV$
\State concatenate heads and apply $\mW_O$
\end{algorithmic}
\end{algorithm}

The matrix has $B\nhead T^2$ entries. For $B=4$, $\nhead=8$, $T=2048$, and float32 storage, this is about $512$ MiB for one attention-probability tensor.

\subsubsection{Design pressure}

Longer contexts are useful only if the $T\times T$ attention matrix is feasible. The arithmetic cost of dense attention remains quadratic, but writing and reading the whole attention matrix from GPU high-bandwidth memory is a major bottleneck.

\subsubsection{Replacement mechanism}

FlashAttention computes the same attention function as \cref{eq:attention}, but avoids materializing the full $T\times T$ matrix in high-bandwidth memory. It uses tiling and an online softmax recurrence to keep only blocks of Q, K, V and running normalization statistics in faster on-chip memory \cite{dao2022flashattention}.

For one query row with score blocks $\vz^{(1)},\vz^{(2)},\ldots$, the online softmax maintains a running maximum $m$ and denominator $\ell$:
\begin{align}
    m_{\mathrm{new}}&=\max(m_{\mathrm{old}},\max_j z^{(b)}_j),\\
    \ell_{\mathrm{new}}&=
    e^{m_{\mathrm{old}}-m_{\mathrm{new}}}\ell_{\mathrm{old}}
    +\sum_j e^{z^{(b)}_j-m_{\mathrm{new}}}.
\end{align}
The partial weighted value sums are rescaled by the same factors, so the final result matches standard softmax attention up to floating-point associativity.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=7mm and 8mm]
    \node[block, minimum width=42mm, fill=orange!10] (hbm) {HBM stores full Q K V tensors};
    \node[block, below=of hbm, minimum width=42mm, fill=blue!8] (load) {load one Q tile and one K/V tile into SRAM};
    \node[block, below=of load, minimum width=42mm, fill=green!8] (local) {compute local scores and update online softmax};
    \node[tensor, below=of local, minimum width=42mm] (out) {write final attention output $\mO$};
    \node[block, right=20mm of local, minimum width=34mm, fill=red!6] (avoid) {never store the full\\$T\times T$ attention matrix};

    \draw[arrow] (hbm) -- (load);
    \draw[arrow] (load) -- (local);
    \draw[arrow] (local) -- (out);
    \draw[grad] (avoid.west) -- (local.east);
\end{tikzpicture}
\caption{FlashAttention motivation. The mathematical attention output is the same, but the computation is tiled so the full attention matrix is not written to GPU high-bandwidth memory.}
\label{fig:flashattention}
\end{figure}

\subsubsection{Trade-off}

FlashAttention is a systems improvement, not a new statistical model. It reduces auxiliary memory and memory traffic, but it does not remove the $O(T^2\dhead)$ arithmetic cost of dense attention. It improves modeling results only indirectly when the saved memory or time is used for longer contexts, larger batches, or more training.


\subsubsection{FlashAttention-2: the same exact function with better GPU work partitioning}

FlashAttention-2 does not change \cref{eq:attention}; it improves how the tiled work is distributed on the GPU \cite{dao2023flashattention2}. The main ideas are:
\begin{enumerate}[nosep]
    \item reduce non-matrix-multiplication operations, which are relatively expensive on GPUs optimized for tensor-core matrix products;
    \item parallelize across the sequence dimension when batch size or head count alone does not create enough independent thread blocks;
    \item reduce communication and synchronization between warps inside a thread block.
\end{enumerate}

\begin{intuitionbox}[Kitchen analogy]
FlashAttention-1 avoids carrying every intermediate dish to a distant warehouse. FlashAttention-2 keeps that good memory strategy but also reorganizes the kitchen staff: more teams can work on one long order, less time is spent handing ingredients between workers, and a larger fraction of time is spent on the high-throughput operation the hardware performs best.
\end{intuitionbox}

For a scientific comparison, the important conceptual rule is to separate \emph{model mathematics} from \emph{kernel efficiency}. If two variants use different kernel families, a wall-clock difference may reflect software dispatch rather than the architectural idea. Validation loss at a fixed token or FLOP budget remains a modeling comparison; elapsed time additionally measures the quality of the available implementation.





\subsection{Sliding-window attention: a local causal inductive bias}
\label{sec:swa}

\subsubsection{Baseline limitation and design pressure}

Full causal attention permits query $i$ to inspect every key $j\leq i$. This is maximally flexible, but the number of allowed query--key pairs grows quadratically:
\begin{equation}
    P_{\mathrm{full}}
    =\sum_{i=1}^{T}i
    =\frac{T(T+1)}{2}.
\end{equation}
Even an IO-efficient exact kernel must perform work for these pairs. Sliding-window attention (SWA) assumes that many transformations, especially in lower layers, can be learned from a local neighborhood \cite{jiang2023mistral}.

\subsubsection{Windowed causal distribution}

With window length $W$, query $i$ may attend only to
\begin{equation}
    \mathcal{J}_i
    =\{j:\max(1,i-W+1)\leq j\leq i\}.
\end{equation}
The logits and probabilities are
\begin{align}
    A_{ij}
    &=\begin{cases}
       \vq_i^\top\vk_j/\sqrt{\dhead}, & j\in\mathcal{J}_i,\\
       -\infty, & j\notin\mathcal{J}_i,
      \end{cases}\\
    \alpha_{ij}
    &=\frac{\exp(A_{ij})}
      {\sum_{u\in\mathcal{J}_i}\exp(A_{iu})},\\
    \vo_i&=\sum_{j\in\mathcal{J}_i}\alpha_{ij}\vv_j.
\end{align}
Here $\mathcal{J}_i$ is the visible set, $\alpha_{ij}$ is a probability distribution over that set, and $\vo_i$ is the weighted value average.

The number of allowed pairs is
\begin{equation}\label{eq:swa-pairs}
    P_{\mathrm{SWA}}
    =\sum_{i=1}^{T}\min(i,W)
    =TW-\frac{W(W-1)}{2}
    \quad (W\leq T).
\end{equation}
For $T\gg W$, this is approximately $TW$ rather than $T^2/2$.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.78]
    \fill[green!16] (0,0) -- (0,5) -- (1.5,5) -- (5,1.5) -- (5,0) -- cycle;
    \fill[red!8] (1.5,5) -- (5,5) -- (5,1.5) -- cycle;
    \draw[step=0.5cm, gray!55] (0,0) grid (5,5);
    \draw[thick] (0,0) rectangle (5,5);
    \draw[very thick, blue!70!black] (0,0) -- (5,5);
    \draw[very thick, dashed, blue!70!black] (0,1.5) -- (3.5,5);
    \node[note] at (1.25,2.4) {visible band};
    \node[note] at (3.9,4.15) {too old or future};
    \node[below] at (2.5,-0.35) {key position $j$};
    \node[rotate=90] at (-0.45,2.5) {query position $i$};
\end{tikzpicture}
\caption{Banded causal mask for sliding-window attention. Each row keeps only the diagonal band containing the current token and its $W-1$ most recent predecessors.}
\label{fig:swa-band}
\end{figure}

\subsubsection{How depth expands the receptive field}

One local layer can move information backward by at most $W-1$ positions. After $L$ consecutive local layers, a path can span at most
\begin{equation}\label{eq:swa-receptive}
    D_L=L(W-1)
\end{equation}
positions, so the number of potentially influential positions is at most $1+D_L$, clipped by the available prefix length. This is a graph-reachability statement, not a guarantee that information is preserved perfectly. Each layer must compress and relay useful distant information through finite-dimensional residual vectors.

\begin{example}[Message relay]
With $W=4$, token 12 cannot directly read token 1 in one layer. It can read as far back as token 9. In a second layer, token 9's representation may already contain information from token 6; after enough layers, a chain of local updates can carry information from token 1 to token 12. This resembles passing a message along neighboring people: reach grows with the number of handoffs, but details may be lost at each handoff.
\end{example}

\subsubsection{Training--generation asymmetry in this project}

A standard serving-oriented SWA design uses the same window at generation and can evict cache entries older than $W$, reducing cache memory to $O(W)$. The project intentionally tests a different hypothesis: local attention during training, but full cached attention during generation. This preserves the full cache and therefore gives no inference-memory saving.

\begin{tradeoffbox}[Why the asymmetry is scientifically interesting but risky]
Full attention at inference offers extra edges the model never used during training. Those distant logits may be poorly calibrated, so additional context can help, do nothing, or even distract. The experiment should compare local-at-train/local-at-test against local-at-train/global-at-test before interpreting any gain as a property of SWA itself.
\end{tradeoffbox}

\subsection{Interleaved local and global attention}
\label{sec:interleaved-swa}

Pure SWA makes every layer relay long-range information indirectly. Interleaving local and global layers asks whether occasional direct global access is enough. Let
\begin{equation}
    \mathcal{A}^{(\ell)}=\begin{cases}
        \text{full causal attention}, & \ell\text{ even},\\
        \text{windowed causal attention}, & \ell\text{ odd}.
    \end{cases}
\end{equation}
This pattern is related to local--global interleaving used in Gemma 2 \cite{gemma2024}.

If half the layers are global and half use window $W$, the allowed-pair count across $L$ layers is approximately
\begin{equation}
    P_{\mathrm{interleaved}}
    =\frac{L}{2}P_{\mathrm{full}}
     +\frac{L}{2}P_{\mathrm{SWA}}.
\end{equation}
The exact saving depends on causal triangular work and the kernel, but the conceptual compromise is clear: local layers spend less computation; global layers periodically restore direct access to the whole prefix.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=5mm and 7mm]
    \node[tensor] (x0) {input};
    \node[block, right=of x0, fill=green!9] (g1) {global\\layer 1};
    \node[block, right=of g1, fill=orange!10] (l2) {local\\layer 2};
    \node[block, right=of l2, fill=green!9] (g3) {global\\layer 3};
    \node[block, right=of g3, fill=orange!10] (l4) {local\\layer 4};
    \node[tensor, right=of l4] (out) {output};
    \draw[arrow] (x0)--(g1); \draw[arrow] (g1)--(l2); \draw[arrow] (l2)--(g3); \draw[arrow] (g3)--(l4); \draw[arrow] (l4)--(out);
    \node[note, below=8mm of l2, xshift=20mm, text width=11cm] {Global layers retrieve distant information directly; local layers perform cheaper neighborhood processing between those retrieval stages.};
\end{tikzpicture}
\caption{Interleaved attention alternates global and local connectivity rather than imposing the same sparsity pattern at every depth.}
\label{fig:interleaved-attn}
\end{figure}

A high-quality result would be evidence that direct global access is needed only periodically. A degradation would suggest either that more global layers are required or that long-range information must be available at particular depths, not merely somewhere in the stack.

\subsection{Linear attention with a positive feature map}
\label{sec:linear-attention}

\subsubsection{From pairwise softmax to a prefix state}

Softmax attention uses the kernel
\begin{equation}
    \kappa_{\mathrm{soft}}(\vq,\vk)
    =\exp(\vq^\top\vk/\sqrt{\dhead}).
\end{equation}
The denominator for query $i$ depends on every previous key, so exact softmax attention naturally performs pairwise work. Linear attention replaces this kernel with an inner product of feature maps,
\begin{equation}
    \kappa(\vq,\vk)=\varphi(\vq)^\top\varphi(\vk),
\end{equation}
where $\varphi:\reals^{\dhead}\to\reals^r$ and $r$ is the feature dimension \cite{katharopoulos2020linear}. This defines a different attention family unless $\varphi$ is specifically constructed as an approximation to the exponential kernel.

For causal attention,
\begin{equation}\label{eq:linear-attn-sum}
    \vo_i=
    \frac{\sum_{j\leq i}\varphi(\vq_i)^\top\varphi(\vk_j)\vv_j}
         {\sum_{j\leq i}\varphi(\vq_i)^\top\varphi(\vk_j)}.
\end{equation}
Associativity allows the query-dependent vector to be moved outside the prefix sums. Define
\begin{align}
    \mS_i&=\sum_{j=1}^{i}\varphi(\vk_j)\vv_j^\top
          =\mS_{i-1}+\varphi(\vk_i)\vv_i^\top,
          &\mS_i&\in\reals^{r\times d_v},\\
    \vz_i&=\sum_{j=1}^{i}\varphi(\vk_j)
          =\vz_{i-1}+\varphi(\vk_i),
          &\vz_i&\in\reals^{r}.
\end{align}
Then
\begin{equation}\label{eq:linear-attn-recurrence}
    \vo_i=
    \frac{\varphi(\vq_i)^\top\mS_i}
         {\varphi(\vq_i)^\top\vz_i+\varepsilon}.
\end{equation}
Here $\mS_i$ is a matrix-valued content summary, $\vz_i$ normalizes total similarity mass, $d_v$ is value dimension, and $\varepsilon>0$ protects finite-precision division.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.82, transform shape, node distance=6mm and 6mm]
    \node[tensor] (state0) {state\\$(\mS_{i-1},\vz_{i-1})$};
    \node[data, below=of state0] (kv) {new $\vk_i,\vv_i$};
    \node[block, right=of state0, yshift=-5mm] (update) {rank-one state update\\$\mS_i=\mS_{i-1}+\varphi(\vk_i)\vv_i^\top$};
    \node[data, right=of update] (q) {query $\vq_i$};
    \node[block, right=of q] (read) {normalized state read\\\cref{eq:linear-attn-recurrence}};
    \node[tensor, right=of read] (o) {output $\vo_i$};
    \draw[arrow] (state0)--(update); \draw[arrow] (kv)--(update); \draw[arrow] (update)--(read); \draw[arrow] (q)--(read); \draw[arrow] (read)--(o);
\end{tikzpicture}
\caption{Causal linear attention as a recurrent memory. The prefix is compressed into a fixed-size state rather than retained as one K/V pair per token.}
\label{fig:linear-state}
\end{figure}

\subsubsection{Why use $\ELU(x)+1$?}

The project uses the element-wise map
\begin{equation}\label{eq:elu-plus-one}
    \varphi(x)=\ELU(x)+1
    =\begin{cases}
        x+1, & x\geq0,\\
        e^x, & x<0.
      \end{cases}
\end{equation}
It is continuous, continuously differentiable at zero, strictly increasing, and strictly positive. Positivity makes every implicit similarity weight non-negative and ensures the denominator is positive in exact arithmetic after at least one key. It can still become numerically tiny when all relevant feature coordinates are close to zero, which motivates $\varepsilon$.

\begin{example}[Outer-product memory]
Suppose $r=d_v=2$, $\varphi(\vk_i)=(2,1)^\top$, and $\vv_i=(3,-1)^\top$. The state update is
\begin{equation}
    \varphi(\vk_i)\vv_i^\top
    =\begin{pmatrix}2\\1\end{pmatrix}
     \begin{pmatrix}3&-1\end{pmatrix}
    =\begin{pmatrix}6&-2\\3&-1\end{pmatrix}.
\end{equation}
A future query does not retrieve this token directly. It reads a weighted combination of all such outer products accumulated in $\mS$.
\end{example}

\subsubsection{Complexity and information trade-off}

For each head, updating $\mS_i$ and reading it costs $O(rd_v)$ per token. With $r=d_v=\dhead$, sequence cost is $O(T\dhead^2)$ and recurrent-state memory is $O(\dhead^2)$, independent of $T$. Full attention costs $O(T^2\dhead)$ and retains a memory that grows with $T$.

The price is prefix compression. Full attention stores separate keys and values, so a later query can form a new weighting over individual past tokens. Linear attention maps the entire prefix into $(\mS_i,\vz_i)$. Different histories can produce the same state, and information that was merged cannot be separated later. This fixed-size-state bottleneck is a more accurate limitation than calling causal linear attention simply a low-rank attention matrix.

\begin{tradeoffbox}[Arithmetic complexity is not wall-clock speed]
A linear asymptotic expression does not guarantee a faster training run. Matrix-multiplication kernels for softmax attention are highly optimized and parallel. A sequential or poorly fused prefix recurrence can have low hardware occupancy and large launch overhead. Both FLOPs and achieved throughput must be measured.
\end{tradeoffbox}

\subsubsection{Position and causality}

The recurrence is causal by construction because $(\mS_i,\vz_i)$ contains only indices $j\leq i$. Therefore the model is order-sensitive even without an explicit positional embedding: changing token order changes every subsequent prefix state. However, causality alone does not provide a rich representation of relative distance. The published linear-RoPE construction applies rotation after the nonlinear feature map in the numerator and leaves the positive normalization denominator unrotated; the exact corrected equation is given in \cref{sec:v5-failure-correction}.

\subsection{V5 numerical failure, literature audit, and correction}
\label{sec:v5-failure-correction}

\subsubsection{Observed failure}

The first corrected causal V5 retraining attempt still failed in all three seeds. The transition was abrupt: ordinary losses and global gradient norms were followed by non-finite values within one logging interval.
\begin{table}[H]
\centering
\caption{First observed non-finite V5 training step. Logging occurred every ten optimizer steps.}
\label{tab:v5-nan-steps}
\begin{tabular}{@{}rrrr@{}}
\toprule
Seed & Last finite step & First non-finite step & Learning rate \\
\midrule
42   & 1620 & 1630 & $2.96\times10^{-4}$ \\
137  & 1690 & 1700 & $2.95\times10^{-4}$ \\
2024 & 1670 & 1680 & $2.96\times10^{-4}$ \\
\bottomrule
\end{tabular}
\end{table}
The preceding gradient norms were approximately $0.65$--$0.85$, but the first saved checkpoints were already entirely non-finite. Because there was no checkpoint or tensor-level trace between the last finite and first non-finite log entries, the precise first activation in those runs cannot be reconstructed.

\subsubsection{Confirmed implementation defects}

The intended feature map was
\begin{equation}
    \varphi(x)=\ELU(x)+1.
\end{equation}
The original implementation expanded it as
\begin{equation}
    \operatorname{where}(x\geq0,\;x+1,\;e^x).
\end{equation}
These expressions are equal in the finite forward pass, but they are not equivalent computational graphs. For a large positive input, approximately $x\geq89$ in float32 or bfloat16, the unused exponential branch overflows. Its backward path can then form $0\cdot\infty$, producing a NaN gradient even though the selected forward branch is $x+1$. A deterministic regression reproduced this at $x=90$ in both precisions. The corrected implementation evaluates the negative branch as $\exp(\min(x,0))$ before selection. It therefore preserves $e^x$ for negative inputs, never evaluates an overflowing exponential for positive inputs, and retains derivative one on the selected positive branch. This branch-safe algebraic form also avoids the cancellation to exactly zero that \texttt{F.elu(x) + 1} can exhibit in the far negative float32 tail.

A second defect expanded the blast radius. The trainer passed a non-finite gradient set directly to \texttt{clip\_grad\_norm\_}. A single NaN makes the global norm and clipping coefficient NaN, so clipping can turn every otherwise finite gradient into NaN before AdamW runs. This explains an entirely corrupted checkpoint without requiring every layer to fail independently.

The positional construction was also not the one derived in RoFormer \cite{su2021roformer}. V5 had applied rotation before the nonlinear map, computing $\varphi(\mR_i\vq_i)$ and $\varphi(\mR_j\vk_j)$ in both numerator and denominator. Since $\varphi$ is nonlinear,
\begin{equation}
    \varphi(\mR_i\vq_i)\neq\mR_i\varphi(\vq_i)
\end{equation}
in general. Thus the implementation was neither the original non-rotary linear Transformer \cite{katharopoulos2020linear} nor the linear-RoPE equation proposed by RoFormer.

\subsubsection{Corrected formulation}

Define positive, unrotated features
\begin{equation}
    \bm{a}_i=\varphi(\vq_i),\qquad \bm{b}_j=\varphi(\vk_j).
\end{equation}
The corrected causal mechanism follows RoFormer equation 19:
\begin{align}
    \bm{n}_i
    &=\bigl(\mR_i\bm{a}_i\bigr)^\top
      \sum_{j\leq i}\bigl(\mR_j\bm{b}_j\bigr)\vv_j^\top,\\
    d_i
    &=\bm{a}_i^\top\sum_{j\leq i}\bm{b}_j,\\
    \vo_i&=\frac{\bm{n}_i}{\max(d_i,\varepsilon)}.
\end{align}
Rotation therefore affects only the numerator. The denominator retains positive unrotated features, as required to preserve a stable normalization term. The chunked implementation maintains separate numerator and denominator prefix states and is still exactly causal.

The implementation correction has four parts:
\begin{enumerate}
    \item use a branch-safe algebraic implementation of $\ELU(x)+1$ that caps the unused exponential input at zero;
    \item apply RoPE after the feature map in the numerator and leave the denominator unrotated;
    \item evaluate feature maps, prefix-state updates, numerator, and denominator reductions in float32 even under bfloat16 autocast; and
    \item measure gradient finiteness before clipping, abort or roll back before the optimizer step, and call clipping with non-finite rejection enabled.
\end{enumerate}
Focused tests now cover the large-positive feature-map backward pass in float32 and bfloat16, equality between the chunked code and a dense implementation of RoFormer equation 19, and the guarantee that one NaN gradient cannot update or contaminate model parameters.

\subsubsection{Completed corrected retraining and evaluation}

After these changes, V5 was retrained from scratch for all three registered seeds. The corrected runs used the same 15,000-step, 983,040,000-token budget as the other main-scale experiments; their checkpoints and logs were written under a distinct \texttt{fixed} name so that the earlier non-finite artifacts cannot be selected accidentally. Every logged training loss and gradient norm remained finite through the final step, and all three final validation evaluations were finite.

\begin{table}[H]
\centering
\caption{Completed main-scale V5 retraining after the numerical and formulation corrections. Throughput is the end-to-end average reported by the trainer.}
\label{tab:v5-fixed-retraining}
\begin{tabular}{@{}rrrrr@{}}
\toprule
Seed & Final train loss & Final validation loss & Time (s) & Tokens/s \\
\midrule
42   & 3.7715 & 3.7565 & 18,519.5 & 53,081 \\
137  & 3.7694 & 3.7545 & 18,112.3 & 54,275 \\
2024 & 3.7700 & 3.7527 & 18,219.7 & 53,955 \\
\bottomrule
\end{tabular}
\end{table}

The mean final validation loss is $3.7546\pm0.0019$, where the uncertainty is the sample standard deviation across seeds. The corresponding in-context-learning decay exponent is $0.6472\pm0.0072$. The narrow spread and three fully finite trajectories show that the confirmed implementation failures were removed end to end; they do not prove that normalized linear attention is unconditionally stable under every scale or hyperparameter setting.

The full 30-checkpoint evaluation was then regenerated with only the corrected V5 artifacts. Its representative-seed V5 losses are 3.7545 at the fixed-data budget, 3.8508 at the common wall-clock budget, and 3.8002 at the common FLOP budget. The earlier near-zero V5 losses were therefore numerical-corruption artifacts, not evidence of exceptional modeling performance. With valid V5 measurements, the FLOPs--validation-loss Pareto front contains the MoE and vanilla variants rather than V5 and vanilla. The corrected V5 MQAR accuracy was zero for all three seeds under the present probe, and its mean stable-rank diagnostic ranged from 4.28 to 4.50 across runs; these diagnostics do not overturn the loss-based conclusion.

The stability repair has a measurable systems cost. The corrected runs averaged approximately 53,770 tokens/s because feature-map evaluation, prefix accumulation, and normalization reductions remain in float32 under bfloat16 autocast. The immediately preceding unstable causal attempt averaged approximately 69,350 tokens/s, so the safer implementation is about 22\% slower in this setup. Those failed runs are not valid quality baselines, but the throughput difference records the price of the higher-precision recurrent path and reinforces the need to report achieved speed rather than asymptotic complexity alone.

\subsubsection{Residual theoretical risk}

The row-sum denominator itself remains the normalized mechanism proposed by Katharopoulos et al. Subsequent analysis proves that normalized kernel attention can have unbounded gradients and reports higher gradient variability for the $\ELU(x)+1$ feature map \cite{qin2022devil}. Qin et al. propose NormAttention, which removes row-sum division and applies RMSNorm after the unnormalized attention output. That is a literature-backed alternative, but it changes the V5 architecture and therefore must be introduced as an explicit new experimental decision rather than folded silently into this repair. The present correction removes the confirmed implementation defects while preserving the intended normalized-attention comparison.

\subsection{Attention entropy as a routing diagnostic}
\label{sec:attention-entropy}

For a query position $i$, layer $\ell$, and head $h$, let
\begin{equation}
    \bm\alpha_i^{(\ell,h)}
    =(\alpha_{i1},\ldots,\alpha_{in_i})
\end{equation}
be the attention probabilities over the $n_i$ visible keys. Shannon entropy is
\begin{equation}\label{eq:attention-entropy}
    H_i^{(\ell,h)}
    =-\sum_{j=1}^{n_i}\alpha_{ij}\log\alpha_{ij}.
\end{equation}
It satisfies
\begin{equation}
    0\leq H_i^{(\ell,h)}\leq\log n_i.
\end{equation}
Entropy zero means all mass is on one key. Maximum entropy means uniform attention. Because $n_i$ changes with position and attention pattern, a normalized version is often easier to compare:
\begin{equation}
    \widetilde H_i^{(\ell,h)}
    =\frac{H_i^{(\ell,h)}}{\log n_i}
    \in[0,1]
    \qquad (n_i>1).
\end{equation}
Another interpretable quantity is the effective support size
\begin{equation}
    n_{\mathrm{eff}}=\exp(H),
\end{equation}
which equals the number of equally weighted keys that would have the same entropy.

\begin{example}[Sharp versus broad routing]
Weights $(1,0,0,0)$ have $H=0$ and $n_{\mathrm{eff}}=1$. Uniform weights $(1/4,1/4,1/4,1/4)$ have $H=\log4$ and $n_{\mathrm{eff}}=4$. Weights $(0.7,0.1,0.1,0.1)$ lie between them, indicating focused but not single-token routing.
\end{example}

Entropy is descriptive, not a quality score. A copying head should be sharp; a summarizing head may be broad. Uniform attention can mean useful aggregation or failure to discriminate, while low entropy can mean precise retrieval or pathological fixation. It should be interpreted together with token distance, head outputs, ablations, and task performance.

Exact entropy requires access to probabilities or equivalent sufficient statistics. Standard fused attention interfaces often return only the final weighted values, so entropy may be unavailable without a special diagnostic path. This is a measurement limitation, not a mathematical impossibility.


% =============================================================================
\section{Feed-Forward Networks: From ReLU/GELU to SwiGLU}
\label{sec:ffn}
% =============================================================================

\subsection{What the FFN does inside a Transformer block}

Attention answers a routing question: which earlier tokens should the current token read from? The feed-forward network answers a different question: after attention has gathered context, what kind of non-linear transformed update is needed to and get us closer to the real function that we are approximating?

For a batch of hidden states $\mX\in\reals^{B\times T\times \dmodel}$, the FFN is applied independently at every position:
\begin{equation}
    \mathrm{FFN}(\mX)_{b,t,:}=g(\mX_{b,t,:}).
\end{equation}
This does \emph{not} mean the FFN ignores context. The vector $\mX_{b,t,:}$ has already been changed by attention, so it contains information gathered from previous tokens. The FFN transforms that contextual vector into new features.

\begin{intuitionbox}[One-token view]
For one token vector $\vx_t$, attention has already created a contextual summary such as ``this token is probably part of a noun phrase'', ``a closing bracket is expected'', or ``the previous word changes the meaning''. The FFN then turns this contextual information into new coordinates in the residual stream. In this sense, attention is communication, while the FFN is local computation after communication.
\end{intuitionbox}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=7mm and 8mm]
    \node[tensor] (x) {contextual\\token vector $\vx_t$};
    \node[block, right=of x] (expand) {expand\\$\dmodel\to\dff$};
    \node[block, right=of expand] (nonlin) {feature tests\\and gates};
    \node[block, right=of nonlin] (compress) {compress\\$\dff\to\dmodel$};
    \node[tensor, right=of compress] (update) {residual\\update};
    \node[op, below=9mm of update] (plus) {$+$};
    \node[tensor, left=of plus] (old) {old\\$\vx_t$};
    \node[tensor, right=of plus] (newx) {new\\$\vx_t$};

    \draw[arrow] (x) -- (expand);
    \draw[arrow] (expand) -- (nonlin);
    \draw[arrow] (nonlin) -- (compress);
    \draw[arrow] (compress) -- (update);
    \draw[arrow] (update) -- (plus);
    \draw[arrow] (old) -- (plus);
    \draw[arrow] (plus) -- (newx);
\end{tikzpicture}
\caption{The FFN writes a token-wise nonlinear update into the residual stream. The update should be expressive, but it must remain scale-compatible with residual addition. This is why FFN design connects to RMSNorm and residual-output initialization.}
\label{fig:ffn-residual-role}
\end{figure}

\subsection{The standard FFN as a collection of feature detectors}

The V0 FFN has the expand-activate-compress form
\begin{equation}\label{eq:standard_ffn}
    \mathrm{FFN}(\vx)=\mW_2\phi(\mW_1\vx),
\end{equation}
where $\mW_1\in\reals^{\dff\times\dmodel}$, $\mW_2\in\reals^{\dmodel\times\dff}$, and typically $\dff=4\dmodel$. If $\vw_i^\top$ is row $i$ of $\mW_1$ and $\vu_i$ is column $i$ of $\mW_2$, then
\begin{equation}
    a_i=\vw_i^\top\vx,
    \qquad
    h_i=\phi(a_i),
    \qquad
    \mathrm{FFN}(\vx)=\sum_{i=1}^{\dff}h_i\vu_i.
\end{equation}
This equation is the clearest way to understand an activation function. Hidden unit $i$ first asks whether the input vector $\vx$ matches a learned direction $\vw_i$. The answer is the pre-activation $a_i$. The activation function turns $a_i$ into a coefficient $h_i$. The output projection then writes $h_i$ times an output direction $\vu_i$ back into the residual stream.

\begin{motivationbox}[What an activation means inside a network]
When we say ``ReLU is zero for negative inputs'', the input is not usually a raw token or pixel. It is a learned score $a_i=\vw_i^\top\vx$. A positive score means ``the current residual-stream vector aligns with this learned feature direction.'' A negative score means ``it points away from this feature direction.'' ReLU turns this into a hard decision: aligned features are allowed to write; anti-aligned features write nothing.
\end{motivationbox}

\begin{example}[A simple hidden-unit example]
Suppose one FFN hidden unit has learned a direction $\vw_i$ that detects whether a contextual token vector looks like ``the start of a list item.'' If $a_i=\vw_i^\top\vx_t=2.0$, then ReLU outputs $2.0$, and the output direction $\vu_i$ is strongly written into the residual stream. If $a_i=-0.4$, ReLU outputs $0$, so this hidden unit contributes no update for that token. The unit is not choosing a token; it is choosing whether one learned feature should contribute to the next representation.
\end{example}

\subsection{Why ReLU helps}

\begin{definition}[Rectified Linear Unit]
\begin{equation}
    \ReLU(x)=\max(0,x)=x\ind[x>0].
\end{equation}
\end{definition}

ReLU helps neural networks for three basic reasons.

\paragraph{1. It makes the network nonlinear.}
Without activations, the composition of two linear maps is still one linear map:
\begin{equation}
    \mW_2(\mW_1\vx)=(\mW_2\mW_1)\vx.
\end{equation}
Stacking many such layers would not create a richer function class. ReLU breaks this collapse because different inputs activate different subsets of hidden units.

\paragraph{2. It creates input-dependent feature selection.}
For one input, hidden units $3,17,40$ might be active; for another input, units $2,9,51$ might be active. Thus the effective linear map changes with the input. A ReLU network is piecewise linear: it behaves like one linear model in one region of representation space and another linear model in another region.

\paragraph{3. It is cheap and sparse.}
ReLU requires only a comparison with zero. Also, inactive hidden units produce exact zeros. This can reduce unnecessary feature writing and was historically one reason ReLU became a strong baseline.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.94, transform shape, node distance=6mm and 7mm]
    \node[tensor] (x) {token vector\\$\vx_t$};
    \node[block, right=of x] (score) {learned score\\$a_i=\vw_i^\top\vx_t$};
    \node[block, right=of score] (relu) {ReLU\\$h_i=\max(0,a_i)$};
    \node[block, right=of relu] (write) {write direction\\$h_i\vu_i$};
    \node[tensor, right=of write] (resid) {residual\\stream update};
    \draw[arrow] (x) -- (score);
    \draw[arrow] (score) -- (relu);
    \draw[arrow] (relu) -- (write);
    \draw[arrow] (write) -- (resid);
    \node[note, below=8mm of score, align=center] {feature detector};
    \node[note, below=8mm of relu, align=center] {hard gate};
    \node[note, below=8mm of write, align=center] {feature contribution};
\end{tikzpicture}
\caption{A ReLU hidden unit inside an FFN. It is a learned feature detector followed by a hard gate and an output direction.}
\label{fig:relu-hidden-unit}
\end{figure}

\subsection{Why ReLU becomes limiting in Transformer FFNs}

The limitations of ReLU are easiest to see from the coefficient $h_i=\ReLU(a_i)$.

\paragraph{1. Negative evidence is erased.}
If $a_i=-2$, this is not weak evidence; it is strong evidence in the opposite direction. ReLU maps both $a_i=-2$ and $a_i=-0.01$ to the same value $0$. If the negative side is useful, the model must spend another hidden unit learning the opposite direction $-\vw_i$.

\paragraph{2. Inactive examples give zero local gradient.}
For $a_i<0$, $\ReLU'(a_i)=0$. During backpropagation, that particular example gives no direct gradient to the feature detector $\vw_i$. A ReLU unit is not guaranteed to be dead forever, because future batches or earlier layers may change $a_i$, but ReLU does create regions where the unit receives no local learning signal.

\paragraph{3. The hard threshold can waste capacity.}
A large Transformer FFN has many hidden units. If many units are inactive for many examples, the model has paid parameters for features that often do not contribute. Sparsity can be useful, but hard sparsity is not always the best use of parameters in a dense language model.

\paragraph{4. Feature selection and feature magnitude are tied.}
The same scalar $a_i$ decides both whether feature $i$ is allowed through and how large the contribution is. The FFN cannot separately learn ``what information is available'' and ``whether this context permits using it.'' This is the direct motivation for gated FFNs.

\begin{warningbox}[Careful wording]
It is too strong to say that ReLU neurons, once inactive, can never recover. The correct statement is local and example-specific: if $a_i<0$ for an example, the derivative through that ReLU is zero for that example. Across different batches and changing upstream representations, the unit may become active again.
\end{warningbox}

\subsection{GELU and SiLU: smoother single-path gates}

GELU and SiLU keep the same single-path FFN architecture but replace the hard ReLU gate by a smooth gate.

\begin{definition}[Gaussian Error Linear Unit]
\begin{equation}\label{eq:gelu}
    \GELU(x)=x\Phi(x),
\end{equation}
where $\Phi$ is the standard normal CDF.
\end{definition}

GELU can be read as ``multiply $x$ by a smooth probability-like measure of being positive.'' For large positive $x$, $\Phi(x)\approx 1$, so $\GELU(x)\approx x$. For large negative $x$, $\Phi(x)\approx 0$, so $\GELU(x)\approx 0$. Near zero, the transition is gradual rather than abrupt. Its derivative is
\begin{equation}
    \GELU'(x)=\Phi(x)+x\phi(x),
\end{equation}
where $\phi$ is the standard normal density. GELU was used in GPT-style models \cite{hendrycks2016gelu,radford2019gpt2}.

\begin{definition}[SiLU / Swish]
\begin{equation}\label{eq:silu}
    \SiLU(x)=x\sigma(x).
\end{equation}
\end{definition}

SiLU uses a logistic sigmoid gate instead of a Gaussian CDF gate. Its derivative is
\begin{equation}\label{eq:silu_derivative}
    \SiLU'(x)=\sigma(x)+x\sigma(x)(1-\sigma(x))
    =\sigma(x)\{1+x(1-\sigma(x))\}.
\end{equation}
SiLU allows small negative values to pass with reduced magnitude. It does not have an entire negative half-line with zero derivative, although the derivative can become very small and has an isolated zero at a negative value.

\begin{example}[Numerical intuition]
For $x=-1$, ReLU gives $0$, while SiLU gives approximately $-0.27$ and GELU gives approximately $-0.16$. Thus smooth activations do not say ``negative means impossible.'' They say ``negative evidence should usually be suppressed, but perhaps not deleted completely.'' For $x=2$, all three activations pass most of the value through, but GELU and SiLU still do so smoothly.
\end{example}

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=0.9cm,y=0.65cm]
    \draw[->, thick] (-4.2,0) -- (4.4,0) node[right] {$x$};
    \draw[->, thick] (0,-1.2) -- (0,4.2) node[above] {$\phi(x)$};
    \draw[very thick, blue!70!black] (-4,0) -- (0,0) -- (4,4);
    \draw[very thick, green!55!black, smooth, domain=-4:4, samples=120] plot (\x,{\x/(1+exp(-\x))});
    \draw[very thick, orange!80!black, smooth, domain=-4:4, samples=120] plot (\x,{0.5*\x*(1+tanh(0.7978845608*(\x+0.044715*\x*\x*\x)))});
    \draw[dashed] (0,0) -- (0,-0.9);
    \node[note, blue!70!black] at (3.45,3.75) {ReLU};
    \node[note, green!55!black] at (3.55,2.55) {SiLU};
    \node[note, orange!80!black] at (3.55,3.15) {GELU};
    \node[note, align=left] at (-2.55,-0.75) {smooth gates keep\\some negative signal};
\end{tikzpicture}
\caption{ReLU, GELU, and SiLU. GELU and SiLU are smoother single-path gates; they improve the activation curve but do not yet separate content from gating.}
\label{fig:relu-gelu-silu-shapes}
\end{figure}

\begin{connectionbox}[Why GELU and SiLU are only an intermediate step]
GELU and SiLU fix the hard-threshold problem, but they still use one projection. The same scalar both represents a candidate feature and decides how much of that feature should pass. SwiGLU changes the architecture by giving these two roles separate learned projections.
\end{connectionbox}

\subsection{SwiGLU: separating content from permission}

A gated FFN explicitly separates the two roles that ReLU combines:
\begin{equation}
    \text{content path: }\vu=\mW_{\mathrm{up}}\vx,
    \qquad
    \text{gate path: }\vg=\SiLU(\mW_{\mathrm{gate}}\vx).
\end{equation}
The SwiGLU FFN is
\begin{equation}\label{eq:swiglu}
    \mathrm{SwiGLU}(\vx)=
    \mW_{\mathrm{down}}\left[\SiLU(\mW_{\mathrm{gate}}\vx)\odot(\mW_{\mathrm{up}}\vx)\right].
\end{equation}
Here
\begin{equation}
    \mW_{\mathrm{gate}},\mW_{\mathrm{up}}\in\reals^{d_h\times\dmodel},
    \qquad
    \mW_{\mathrm{down}}\in\reals^{\dmodel\times d_h}.
\end{equation}
SwiGLU was proposed as one of several GLU variants for Transformer FFNs \cite{shazeer2020glu} and is used in LLaMA-style architectures \cite{touvron2023llama}.

\begin{motivationbox}[Why the project swaps ReLU/GELU to SwiGLU]
The project swaps to SwiGLU because the FFN should do more than ask whether one learned scalar is positive. It should learn conditional feature computation: ``this content is available, and this context decides how much of it should be written.'' The up path supplies candidate content; the gate path supplies context-dependent permission; their product performs feature-wise selection.
\end{motivationbox}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=8mm and 9mm]
    \node[tensor] (x1) {$\vx$};
    \node[block, right=of x1] (w1) {$\mW_1$};
    \node[block, right=of w1] (relu) {ReLU/GELU/SiLU\\single path};
    \node[block, right=of relu] (w2) {$\mW_2$};
    \node[tensor, right=of w2] (out1) {update};

    \node[tensor, below=20mm of x1] (x2) {$\vx$};
    \node[block, above right=2mm and 10mm of x2] (wg) {$\mW_{\mathrm{gate}}$};
    \node[block, right=of wg] (sg) {SiLU};
    \node[block, below right=2mm and 10mm of x2] (wu) {$\mW_{\mathrm{up}}$};
    \node[op, right=32mm of x2] (mul) {$\odot$};
    \node[block, right=of mul] (wd) {$\mW_{\mathrm{down}}$};
    \node[tensor, right=of wd] (out2) {update};

    \draw[arrow] (x1) -- (w1);
    \draw[arrow] (w1) -- (relu);
    \draw[arrow] (relu) -- (w2);
    \draw[arrow] (w2) -- (out1);

    \draw[arrow] (x2) -- (wg);
    \draw[arrow] (wg) -- (sg);
    \draw[arrow] (sg) -- (mul);
    \draw[arrow] (x2) -- (wu);
    \draw[arrow] (wu) -- (mul);
    \draw[arrow] (mul) -- (wd);
    \draw[arrow] (wd) -- (out2);

    \node[note, above=2mm of relu] {old: one scalar does both jobs};
    \node[note, below=1mm of wu] {content path};
    \node[note, above=1mm of sg] {gate path};
\end{tikzpicture}
\caption{The core ReLU-to-SwiGLU shift. ReLU/GELU/SiLU use one hidden path; SwiGLU separates candidate content from the gate that decides how much content to write.}
\label{fig:relu-to-swiglu}
\end{figure}

\begin{example}[A simple gate-content example]
Suppose the content path proposes $u_i=5$, meaning ``write a strong feature in direction $\vu_i$.'' If the gate is $g_i=0.05$, the product is $0.25$, so the feature is almost suppressed. If the gate is $g_i=1.8$, the product is $9$, so the feature is amplified. The important point is that the model can learn content and permission separately: the content may be useful, but only under certain contexts.
\end{example}

\subsection{Advantages of SwiGLU over ReLU/GELU}

\paragraph{1. Smooth gating.}
The gate uses SiLU, so the transition from suppressed to active is smooth rather than a hard half-line cutoff.

\paragraph{2. Separate content and selection.}
The up projection can specialize in candidate features, while the gate projection can specialize in deciding which candidate features matter for the current context.

\paragraph{3. Multiplicative interactions.}
The product $\vg\odot\vu$ creates conditional interactions between two learned projections. A plain ReLU FFN can approximate such effects, but SwiGLU represents them more directly.

\paragraph{4. Parameter-efficient quality.}
When the hidden dimension is adjusted so that parameter count is comparable, gated variants such as SwiGLU have empirically improved Transformer FFNs over ReLU/GELU baselines \cite{shazeer2020glu}. The fair comparison is parameter-matched, not width-matched.

\paragraph{5. Good fit with pre-normalized residual blocks.}
RMSNorm controls the scale of $\vx$ before the gate and content projections. Residual-output initialization controls the scale of $\mW_{\mathrm{down}}$ when writing the gated update back into the residual stream. Thus SwiGLU is not isolated; it relies on the surrounding architecture to keep multiplicative updates stable.

\subsection{Costs introduced by SwiGLU}

\begin{tradeoffbox}[SwiGLU is not free]
SwiGLU improves expressivity, but it introduces costs.
\begin{enumerate}[nosep]
    \item \textbf{Extra projection:} ReLU/GELU FFNs use two matrices; SwiGLU uses three.
    \item \textbf{Activation memory:} the gate and content activations may both be stored for backpropagation.
    \item \textbf{Less exact sparsity:} ReLU gives exact zeros; SiLU gates are usually nonzero.
    \item \textbf{Scale sensitivity:} multiplying two learned projections can increase activation variance if not controlled.
    \item \textbf{More design choices:} the hidden dimension $d_h$ must be chosen and often rounded to a hardware-friendly multiple.
\end{enumerate}
\end{tradeoffbox}

\subsection{Parameter matching: why the hidden width changes}

The standard FFN with $\dff=4\dmodel$ has
\begin{equation}
    2\dmodel\dff=8\dmodel^2
\end{equation}
parameters. SwiGLU has three matrices:
\begin{equation}
    3\dmodel d_h.
\end{equation}
Matching total parameter count gives
\begin{equation}
    3\dmodel d_h=8\dmodel^2
    \qquad\Rightarrow\qquad
    d_h=\frac{8}{3}\dmodel.
\end{equation}
In code, $d_h$ is often rounded to a multiple such as $64$ or $256$ for efficient tensor-core use.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=9mm and 8mm]
    \node[block] (relu) {ReLU/GELU FFN\\$\dmodel\to4\dmodel\to\dmodel$};
    \node[block, right=22mm of relu] (swiglu) {SwiGLU FFN\\two $\dmodel\to d_h$ paths\\one $d_h\to\dmodel$ path};
    \node[note, below=8mm of relu] (p1) {$2\dmodel(4\dmodel)=8\dmodel^2$};
    \node[note, below=8mm of swiglu] (p2) {$3\dmodel d_h$};
    \node[op, below=19mm of relu, xshift=34mm] (eq) {$=$};
    \node[note, below=9mm of eq] {$d_h=\frac{8}{3}\dmodel$};
    \draw[arrow] (p1) -- (eq);
    \draw[arrow] (p2) -- (eq);
\end{tikzpicture}
\caption{Parameter matching for ReLU/GELU FFN versus SwiGLU FFN. SwiGLU spends parameters on a separate gate path, so the hidden width is reduced from $4\dmodel$ to about $(8/3)\dmodel$.}
\label{fig:swiglu-param-matching}
\end{figure}

\begin{connectionbox}[One-sentence summary]
ReLU is a hard one-path feature detector; GELU and SiLU are smoother one-path gates; SwiGLU is a two-path gated FFN that separates candidate content from context-dependent permission, at the cost of another projection and more activation memory.
\end{connectionbox}

\subsection{Mixture of Experts: capacity through conditional computation}
\label{sec:moe}

\subsubsection{Why a dense FFN couples capacity and cost}

In a dense Transformer layer, every token uses the same SwiGLU FFN and therefore activates every FFN parameter. Widening the FFN increases both total model capacity and per-token computation. Mixture of Experts (MoE) breaks this coupling: it stores many expert FFNs but routes each token through only a small subset \cite{fedus2022switch,jiang2024mixtral}.

The central hypothesis is that tokens do not all require the same transformation. A token inside source code, a proper noun, a punctuation mark, and a mathematical operator may benefit from different nonlinear feature banks. Sparse routing allows experts to specialize while keeping the active computation per token bounded.

\subsubsection{Router, top-$k$ selection, and weighted expert output}

Let $N_{\mathrm{exp}}$ be the number of experts. For token representation $\vx\in\reals^{\dmodel}$, a router produces logits
\begin{equation}
    \vz=\mW_r\vx\in\reals^{N_{\mathrm{exp}}},
\end{equation}
where $\mW_r\in\reals^{N_{\mathrm{exp}}\times\dmodel}$. Softmax probabilities are
\begin{equation}
    p_i=\frac{e^{z_i}}{\sum_{j=1}^{N_{\mathrm{exp}}}e^{z_j}},
    \qquad i=1,\ldots,N_{\mathrm{exp}}.
\end{equation}
Let $\mathcal{T}(\vx)$ contain the indices of the $k$ largest probabilities. The selected probabilities are renormalized,
\begin{equation}
    \widetilde p_i
    =\frac{p_i}{\sum_{j\in\mathcal{T}(\vx)}p_j},
    \qquad i\in\mathcal{T}(\vx),
\end{equation}
and the MoE output is
\begin{equation}\label{eq:moe-output-full}
    \mathrm{MoE}(\vx)
    =\sum_{i\in\mathcal{T}(\vx)}
      \widetilde p_i\,E_i(\vx),
\end{equation}
where $E_i$ is expert $i$, usually an independent SwiGLU FFN.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=6mm and 8mm]
    \node[tensor] (x) {token state $\vx$};
    \node[block, right=of x] (router) {router\\probabilities $\bm p$};
    \node[smallblock, above right=5mm and 13mm of router] (e1) {expert 1};
    \node[smallblock, right=13mm of router] (e2) {expert 2};
    \node[smallblock, below right=5mm and 13mm of router] (e3) {expert 3};
    \node[smallblock, below=12mm of e3] (eN) {expert $N_{\mathrm{exp}}$};
    \node[block, right=28mm of router] (mix) {weighted sum of\\top-$k$ outputs};
    \node[tensor, right=of mix] (out) {residual update};
    \draw[arrow] (x)--(router);
    \draw[arrow] (router)--node[above, note] {$\widetilde p_1$}(e1);
    \draw[arrow] (router)--node[above, note] {$\widetilde p_2$}(e2);
    \draw[dashed, gray, ->] (router)--(e3);
    \draw[dashed, gray, ->] (router)--(eN);
    \draw[arrow] (e1)--(mix); \draw[arrow] (e2)--(mix); \draw[arrow] (mix)--(out);
    \node[note, below=3mm of eN] {unselected experts do no token-level FFN work};
\end{tikzpicture}
\caption{Top-$2$ sparse routing. The router selects two experts, computes only their FFN outputs, and mixes them using renormalized routing weights.}
\label{fig:moe-routing}
\end{figure}

\begin{example}[Specialist workshop]
A large workshop owns eight specialist machines, but each job visits only two. A metal part may use cutting and polishing; a printed circuit may use soldering and inspection. The workshop's total capability grows with all eight machines, while the time for one job depends mainly on the two machines it actually uses. MoE applies the same capacity-versus-active-cost idea to token transformations.
\end{example}

\subsubsection{Total parameters versus active parameters}

Suppose one expert contains $P_E$ parameters and shared non-expert components contain $P_S$. An all-MoE stack has approximately
\begin{align}
    P_{\mathrm{total}}&=P_S+L_{\mathrm{MoE}}N_{\mathrm{exp}}P_E,\\
    P_{\mathrm{active/token}}&=P_S^{\mathrm{active}}+L_{\mathrm{MoE}}kP_E.
\end{align}
The distinction matters: total parameters determine storage capacity and memory footprint; active parameters are more closely related to per-token expert FLOPs. Their ratio is only approximately $N_{\mathrm{exp}}/k$ because embeddings, attention, routers, and any dense layers are shared.

This is why parameter parity must be defined carefully for MoE. Matching V6 to a dense model by \emph{active} parameters controls per-token capacity better than matching total parameters, but total memory remains much larger.

\subsubsection{Why routers collapse without balancing pressure}

A router may discover that one expert is slightly better early in training. More tokens then visit that expert, giving it more gradient updates, making it even better, and creating a positive feedback loop. This is expert collapse.

For a batch containing $M=BT$ token positions, define
\begin{align}
    f_i&=\frac{1}{M}\sum_{n=1}^{M}
         \ind[i\in\mathcal{T}(\vx_n)]/k,\\
    P_i&=\frac{1}{M}\sum_{n=1}^{M}p_i(\vx_n).
\end{align}
Here $f_i$ is the realized fraction of selected expert slots assigned to expert $i$, while $P_i$ is its average soft routing probability. A common auxiliary loss is
\begin{equation}\label{eq:moe-balance}
    \loss_{\mathrm{bal}}
    =\alpha N_{\mathrm{exp}}
      \sum_{i=1}^{N_{\mathrm{exp}}}f_iP_i,
\end{equation}
where $\alpha\geq0$ controls the strength. The hard counts $f_i$ are treated as constants during backpropagation; gradients flow through $P_i$. If an expert is overused, its large $f_i$ makes extra probability mass on it expensive.

\begin{warningbox}[Uniform routing is not the final goal]
The balancing loss prevents catastrophic concentration; it should not force experts to be identical or used exactly equally for every data domain. Too large an $\alpha$ can override the language-model objective and punish useful specialization.
\end{warningbox}

Router logits can also grow to unnecessarily large magnitudes, making softmax nearly discrete and optimization brittle. A router z-loss penalizes the squared log-partition function:
\begin{equation}\label{eq:moe-zloss}
    \loss_z
    =\beta\frac{1}{M}
      \sum_{n=1}^{M}
      \left[\log\left(\sum_{i=1}^{N_{\mathrm{exp}}}e^{z_{n,i}}\right)\right]^2,
\end{equation}
where $\beta\geq0$ is a coefficient and $z_{n,i}$ is the router logit for token $n$ and expert $i$ \cite{zoph2022stmoe}. The training loss becomes
\begin{equation}
    \loss_{\mathrm{total}}
    =\loss_{\mathrm{CE}}+\loss_{\mathrm{bal}}+\loss_z.
\end{equation}

\subsubsection{Where to place expert layers}

Three placements answer different scientific questions:
\begin{itemize}[nosep]
    \item \textbf{all layers:} maximum sparse capacity, maximum expert memory;
    \item \textbf{interleaved dense/MoE:} tests whether specialization is needed at every depth;
    \item \textbf{deep-only MoE:} tests whether early layers learn general lexical/syntactic features while deeper layers benefit more from specialized transformations.
\end{itemize}

\subsubsection{New trade-offs}

MoE increases total capacity without proportional expert FLOPs, but introduces memory, routing imbalance, communication, and statistical-efficiency problems. With limited data, experts may receive too few examples or converge to redundant functions. On a single GPU, all expert weights must still fit in memory. In distributed systems, routing tokens to experts can create expensive all-to-all communication. Therefore V6 should be evaluated on loss, active FLOPs, total memory, expert utilization, and expert diversity rather than on parameter count alone.

% =============================================================================
\section{Autoregressive Generation and Decoding}
\label{sec:generation}
% =============================================================================

Training and generation use the same conditional distribution,
\begin{equation}
    p_{\vtheta}(s_{t+1}\given s_{1:t}),
\end{equation}
but they use it differently. Training evaluates many positions in parallel and compares them with known next tokens. Generation has no future tokens available; it repeatedly samples or selects one token, appends it to the context, and runs the model again.

\subsection{Why naive generation repeats work}

At generation step $t$, a naive implementation feeds the whole prefix $s_{1:t}$ through every layer again. This recomputes hidden states, keys, and values for all earlier positions, even though those earlier tokens have not changed. The waste is easiest to see from causal attention: the key and value at position $j$ depend only on $s_{\leq j}$, because the causal mask prevents dependence on future tokens. Therefore, once $\vk_j$ and $\vv_j$ are computed, future tokens cannot invalidate them.

\begin{intuitionbox}[Caching in ordinary language]
If you have already summarized chapter 1 of a book, you do not need to re-summarize chapter 1 from scratch every time you read a new sentence in chapter 2. You keep the old summary and add the new information. The KV-cache is the Transformer's version of keeping old summaries: it stores old keys and values so the new query can read from them.
\end{intuitionbox}

\subsection{KV-cache definition and mechanism}

\begin{definition}[KV-cache]
For each layer $\ell$, the KV-cache stores the keys and values already computed for previous positions:
\begin{equation}
    \mathrm{cache}^{(\ell)}_t=\left(\mK^{(\ell)}_{1:t},\mV^{(\ell)}_{1:t}\right).
\end{equation}
At step $t+1$, the model computes only $\vq_{t+1},\vk_{t+1},\vv_{t+1}$ for the new token, appends $\vk_{t+1},\vv_{t+1}$ to the cache, and attends the new query to all cached keys and values.
\end{definition}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=7mm and 8mm]
    \node[data] (new) {new token\\$s_t$};
    \node[block, right=of new] (proj) {compute\\$\vq_t,\vk_t,\vv_t$};
    \node[block, right=of proj] (append) {append\\$\vk_t,\vv_t$};
    \node[tensor, above=9mm of append, minimum width=38mm] (cache) {cached keys and values\\$\mK_{1:t-1},\mV_{1:t-1}$};
    \node[block, right=of append] (attn) {attend\\$\vq_t$ to cache};
    \node[tensor, right=of attn] (out) {next-token\\logits};

    \draw[arrow] (new) -- (proj);
    \draw[arrow] (proj) -- (append);
    \draw[arrow] (cache) -- (append);
    \draw[arrow] (append) -- (attn);
    \draw[arrow] (cache) -| (attn);
    \draw[arrow] (attn) -- (out);
    \node[note, below=6mm of append] {past K/V tensors are reused rather than recomputed};
\end{tikzpicture}
\caption{KV-cache during autoregressive generation. The new query reads from stored keys and values from previous positions.}
\label{fig:kv-cache}
\end{figure}

\subsection{Compute, memory, and limitations of the KV-cache}

The KV-cache does not remove the need to attend over previous tokens. The new query still compares itself with $t$ cached keys. The gain is that the model does not recompute old hidden states and old K/V projections.

The cache memory across all layers is
\begin{equation}\label{eq:kvcache_memory}
    2\cdot\nlayer\cdot B\cdot\nhead\cdot T\cdot\dhead\cdot s,
\end{equation}
where the factor $2$ is for K and V, and $s$ is bytes per element. Since $\nhead\dhead=\dmodel$ for standard MHA, this is also
\begin{equation}
    2\cdot\nlayer\cdot B\cdot T\cdot\dmodel\cdot s.
\end{equation}

\begin{example}[Cache size]
For $\nlayer=8$, $B=1$, $\nhead=8$, $T=1024$, $\dhead=64$, and bf16 ($s=2$ bytes), the cache uses
\begin{equation}
    2\cdot8\cdot1\cdot8\cdot1024\cdot64\cdot2=16\text{ MiB}.
\end{equation}
This is small for one sequence, but it grows linearly with batch size, generated length, and number of simultaneous users.
\end{example}

\begin{tradeoffbox}[KV-cache trade-off]
The KV-cache changes inference computation but not the learned distribution. It buys speed by spending memory. Long-context serving is often limited by cache memory rather than by the parameter memory of the model itself.
\end{tradeoffbox}

\subsection{KV-cache with RoPE}

With RoPE, keys are stored after rotation:
\begin{equation}
    \vk_t=\mW_K\vx_t,
    \qquad
    \widetilde{\vk}_t=\mR_t\vk_t.
\end{equation}
The cache stores $\widetilde{\vk}_t$. Future queries are rotated by their own positions and compared against already rotated keys. No old key needs to be re-rotated.

\begin{connectionbox}[Why RoPE and caching fit together]
RoPE applies a deterministic position-dependent rotation to each key at creation time. Since causal attention makes old keys independent of future tokens, the rotated key remains valid forever. This is why the correct generation order is: project K, rotate K by its position, then store it in the cache.
\end{connectionbox}

\subsection{KV-cache with grouped-query attention}

The cache formula in \cref{eq:kvcache_memory} assumes one K/V head for every query head. GQA changes only the head-count factor:
\begin{equation}
    \mathrm{cache}_{\mathrm{GQA}}
    =2\nlayer B n_{\mathrm{kv}}T\dhead s.
\end{equation}
The reduction is especially important during token-by-token decoding because each new query must read the cache from memory. Fewer K/V heads reduce both storage and memory bandwidth. The full mechanism, parameter formulas, and MHA--GQA--MQA spectrum are explained once in \cref{sec:gqa}.


\subsection{From logits to a generated token}

After the model produces logits $\vz\in\reals^{\vocab}$, decoding converts those scores into the next token. This is not part of training; it is a decision rule applied to the trained probability distribution. Different decoding rules can make the same model look deterministic, creative, repetitive, or noisy.

\begin{example}[Toy next-token distribution]
Suppose the model gives the following probabilities after the prompt ``The cat sat on the'':
\begin{center}
\begin{tabular}{lcccccc}
\toprule
Token & mat & floor & sofa & roof & banana & quantum\\
\midrule
Probability & 0.42 & 0.24 & 0.16 & 0.08 & 0.06 & 0.04\\
\bottomrule
\end{tabular}
\end{center}
Greedy decoding always picks ``mat''. Sampling may pick ``floor'' or ``sofa'' sometimes. Top-$k$ and top-$p$ decide which low-probability tokens are allowed into the sampling set.
\end{example}

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=1.05cm,y=4.8cm]
    \draw[->, thick] (0,0) -- (7,0) node[right] {tokens sorted by probability};
    \draw[->, thick] (0,0) -- (0,0.55) node[above] {prob.};
    \foreach \x/\h/\lab in {0.6/0.42/mat,1.5/0.24/floor,2.4/0.16/sofa,3.3/0.08/roof,4.2/0.06/banana,5.1/0.04/quantum}{
        \draw[fill=blue!15, thick] (\x,0) rectangle +(0.55,\h);
        \node[rotate=35, anchor=east, font=\scriptsize] at (\x+0.35,-0.015) {\lab};
    }
    \draw[decorate,decoration={brace,amplitude=4pt},thick] (0.55,0.49) -- node[above=4pt, note] {top-$k$ with $k=3$ keeps first three} (2.95,0.49);
    \draw[dashed, red!70!black, thick] (2.95,0) -- (2.95,0.47);
    \node[note, align=center] at (4.6,0.39) {top-$p$ keeps as many\\tokens as needed to pass\\a probability threshold};
\end{tikzpicture}
\caption{Decoding filters operate on the probability distribution after the model has been trained. They change generation behavior but not the model parameters.}
\label{fig:decoding-filters}
\end{figure}

\subsection{Temperature scaling}

Temperature rescales logits before softmax:
\begin{equation}
    p_i(\tau)=\frac{\exp(z_i/\tau)}{\sum_j\exp(z_j/\tau)}.
\end{equation}
For $\tau<1$, the distribution becomes sharper. For $\tau>1$, the distribution becomes flatter. As $\tau\to0$, sampling approaches greedy decoding.

\begin{intuitionbox}[Temperature as confidence control]
Temperature does not change the ranking of tokens; it changes how strongly the model prefers high-ranked tokens. Low temperature says ``trust the model's favorite choices strongly.'' High temperature says ``allow more alternatives.'' Very high temperature can make output incoherent because low-probability tokens become too likely.
\end{intuitionbox}

\subsection{Greedy decoding}

Greedy decoding chooses the most likely next token:
\begin{equation}
    s_{t+1}=\argmax_{w} p_{\vtheta}(w\given s_{\leq t}).
\end{equation}
It is deterministic and fast. Its limitation is that local probability maximization can lead to repetitive or bland text. The most likely next token at each step is not necessarily part of the best full sequence.

\subsection{Top-$k$ sampling}

Top-$k$ keeps only the $k$ highest-probability tokens:
\begin{equation}
    p_{\mathrm{top}\text{-}k}(w)=\begin{cases}
        p(w)/\sum_{u\in S_k}p(u), & w\in S_k,\\
        0, & w\notin S_k,
    \end{cases}
\end{equation}
where $S_k$ is the set of $k$ highest-probability tokens.

\begin{example}[Top-$k$ on the toy distribution]
With $k=3$, the allowed tokens are ``mat'', ``floor'', and ``sofa''. The other tokens receive probability zero. The remaining probabilities are renormalized: the new probability of ``mat'' is $0.42/(0.42+0.24+0.16)\approx0.51$.
\end{example}

The limitation is that $k$ is fixed. If the model is very confident, $k=50$ may include many implausible tokens. If the model is uncertain, $k=50$ may exclude reasonable options.

\subsection{Nucleus sampling or top-$p$ sampling}

Nucleus sampling sorts tokens by descending probability and chooses the smallest prefix $S_p$ whose cumulative probability is at least $p$ \cite{holtzman2019nucleus}:
\begin{equation}
    S_p=\min\left\{S:\sum_{w\in S}p(w)\geq p\right\},
\end{equation}
where the minimum is over prefixes of the sorted token list. Then the distribution is renormalized over $S_p$.

\begin{example}[Top-$p$ on the toy distribution]
If $p=0.8$, we include ``mat'' ($0.42$), ``floor'' ($0.66$ cumulative), and ``sofa'' ($0.82$ cumulative). We stop after ``sofa'' because the cumulative probability has passed $0.8$. If the distribution were flatter, top-$p$ would automatically include more tokens.
\end{example}

\subsection{Decoding is not training}

\begin{motivationbox}[Separate the learned distribution from the decision rule]
Training estimates conditional probabilities by minimizing cross-entropy. Decoding chooses how to use those probabilities. A model can have a lower validation loss but still produce poor samples under a bad decoding rule. Conversely, careful temperature and top-$p$ settings can make a fixed model sound much better without changing its learned parameters.
\end{motivationbox}

Common sampling hyperparameters should therefore be reported with generation examples: temperature, top-$k$, top-$p$, maximum length, repetition penalties if used, and the prompt format.


% =============================================================================
\section{Architectural Simplifications}
\label{sec:simplifications}
% =============================================================================

This section contains smaller architecture choices from the original notes. They should not be re-explained elsewhere: if a future variant changes one of them, add the explanation here.

\subsection{Weight tying}

\begin{definition}[Weight tying]
The input embedding matrix and output classifier matrix share parameters:
\begin{equation}
    \mW_{\mathrm{head}}=\mW_{\mathrm{tok}}.
\end{equation}
\end{definition}
The logit for token $w$ becomes
\begin{equation}
    z_w=\vh^\top\mW_{\mathrm{tok}}[w].
\end{equation}
Thus the model uses the same vector space for reading tokens and predicting tokens. If the final hidden vector $\vh$ points in a similar direction to the embedding of token $w$, then token $w$ receives a large logit.

\begin{example}[Why tying saves parameters]
If $\vocab=50{,}000$ and $\dmodel=512$, the embedding matrix has about $25.6$ million parameters. Without weight tying, the output head would add another $25.6$ million parameters. Tying shares these parameters and also encourages a consistent geometry between input tokens and output predictions.
\end{example}

Weight tying can significantly reduce parameters when $\vocab\dmodel$ is large \cite{press2017weighttying}. Its limitation is that it constrains the input and output token spaces to use the same embedding geometry. In language modeling this is usually beneficial, but it is still a modeling assumption.

\subsection{Mostly bias-free layers}

A standard linear layer uses $\mW\vx+\bm b$. Modern decoder-only LLMs often omit biases in most projections:
\begin{equation}
    \vy=\mW\vx.
\end{equation}
The motivation is practical, not mystical. Bias vectors are small compared with large matrices, and pre-normalization already controls the distribution entering each projection. For Q/K projections with RoPE, bias terms are especially awkward: a fixed vector added before rotation becomes a position-dependent rotated offset.

\begin{connectionbox}[Connection to RMSNorm and RoPE]
RMSNorm uses scale without an additive shift. Bias-free projections extend the same minimalist style. RoPE wants position to enter through rotations of Q and K; removing Q/K biases keeps that geometry cleaner.
\end{connectionbox}

The trade-off is reduced affine flexibility. Bias-free layers are a modern LLM convention, not a theorem that biases are never useful. Small-data fine-tuning or non-language tasks may benefit from biases.

\subsection{No-dropout pretraining}

\begin{definition}[Dropout]
During training, dropout sets activations to zero with probability $p$ and rescales surviving activations:
\begin{equation}
    \mathrm{Dropout}(x_i)=\begin{cases}
        0, & \text{with probability }p,\\
        x_i/(1-p), & \text{with probability }1-p.
    \end{cases}
\end{equation}
\end{definition}

Dropout is a regularizer. It is useful when a model can overfit a limited dataset by relying on fragile co-adapted features. Large-scale pretraining often uses zero dropout because the limiting factor is usually not classical small-data overfitting; it is efficient use of data, compute, and optimization.

\begin{tradeoffbox}[When no dropout can fail]
No-dropout pretraining is not a universal rule. If the training loss keeps decreasing while validation loss increases, the model is overfitting and dropout or another regularizer may help. For fair V0/V1 comparisons, dropout should be fixed across variants unless dropout is the object of the experiment.
\end{tradeoffbox}

% =============================================================================
\section{Training, Numerical Precision, and Evaluation}
\label{sec:training-eval}
% =============================================================================

\subsection{AdamW and weight-decay groups}

Adam keeps moving averages of gradients and squared gradients:
\begin{align}
    \bm m_t&=\beta_1\bm m_{t-1}+(1-\beta_1)\bm g_t,\\
    \bm v_t&=\beta_2\bm v_{t-1}+(1-\beta_2)\bm g_t^2.
\end{align}
Bias-corrected estimates are
\begin{equation}
    \widehat{\bm m}_t=\frac{\bm m_t}{1-\beta_1^t},
    \qquad
    \widehat{\bm v}_t=\frac{\bm v_t}{1-\beta_2^t}.
\end{equation}
AdamW decouples weight decay from the adaptive gradient scaling \cite{loshchilov2019adamw}:
\begin{equation}\label{eq:adamw}
    \vtheta_t=\vtheta_{t-1}
    -\eta_t\frac{\widehat{\bm m}_t}{\sqrt{\widehat{\bm v}_t}+\epsilon}
    -\eta_t\lambda\vtheta_{t-1}.
\end{equation}
A common rule is to apply weight decay to two-dimensional weight matrices, but not to normalization scales, embeddings, or bias vectors.

\subsection{Learning-rate warmup and cosine decay}

A standard schedule is
\begin{equation}\label{eq:cosine_schedule}
    \eta(t)=\begin{cases}
        \eta_{\max}\,t/t_{\mathrm{warmup}}, & t<t_{\mathrm{warmup}},\\[0.5em]
        \eta_{\min}+\dfrac{\eta_{\max}-\eta_{\min}}{2}
        \left(1+\cos\left(\pi\dfrac{t-t_{\mathrm{warmup}}}{t_{\mathrm{total}}-t_{\mathrm{warmup}}}\right)\right), & t\geq t_{\mathrm{warmup}}.
    \end{cases}
\end{equation}
Warmup avoids unstable early updates before Adam's moment estimates are calibrated. Decay reduces optimization noise later.

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=1cm,y=1cm]
    \draw[->, thick] (0,0) -- (7.2,0) node[right] {training step};
    \draw[->, thick] (0,0) -- (0,2.6) node[above] {learning rate $\eta$};
    \draw[very thick, blue!70!black] (0,0) -- (1.2,2.1);
    \draw[very thick, blue!70!black] plot[smooth] coordinates {(1.2,2.1) (2.2,2.0) (3.4,1.55) (4.8,0.85) (6.8,0.25)};
    \draw[dashed] (1.2,0) -- (1.2,2.1);
    \node[note] at (0.6,-0.35) {warmup};
    \node[note] at (4.0,-0.35) {cosine decay};
    \node[note, align=left] at (4.3,2.25) {large steps early,\\smaller steps near convergence};
\end{tikzpicture}
\caption{Learning-rate warmup and decay. The schedule is part of a fair architecture comparison because a bad schedule can hide the value of an architecture.}
\label{fig:lr-schedule}
\end{figure}

\subsection{Gradient clipping and accumulation}

Global norm clipping rescales the full gradient vector if it exceeds threshold $c$:
\begin{equation}\label{eq:grad_clip}
    \bm g\leftarrow \bm g\cdot\min\left(1,\frac{c}{\Vert\bm g\Vert_2}\right).
\end{equation}
This preserves direction when clipping occurs and protects against rare destructive updates.

If the desired effective batch size does not fit in memory, use $A$ micro-batches:
\begin{equation}
    \bm g_{\mathrm{eff}}=\frac{1}{A}\sum_{a=1}^{A}\bm g^{(a)}.
\end{equation}
The optimizer steps once after accumulating $A$ gradients. Usually each micro-batch loss is divided by $A$ before backpropagation.

\subsection{Mixed precision and bfloat16}

\begin{center}
\begin{tabular}{lcccc}
\toprule
Format & Sign bits & Exponent bits & Mantissa bits & Total bits\\
\midrule
float32 & 1 & 8 & 23 & 32\\
float16 & 1 & 5 & 10 & 16\\
bfloat16 & 1 & 8 & 7 & 16\\
\bottomrule
\end{tabular}
\end{center}

bfloat16 has the same exponent width as float32, so it has a much larger dynamic range than float16. Its mantissa is shorter, so individual numbers are less precise, but neural-network gradients are already noisy because they are estimated from mini-batches. In practice, the larger exponent range is often more important than extra mantissa bits.

\begin{intuitionbox}[Why bf16 works]
Mini-batch training already uses an approximate gradient. If the stochastic error from sampling a batch is larger than the rounding error from bf16 arithmetic, then bf16 matrix multiplications can be a good trade-off. However, reductions such as softmax denominators, normalization statistics, and losses are more sensitive to accumulation error, so they are often kept in float32.
\end{intuitionbox}

A typical mixed-precision strategy is:
\begin{itemize}[nosep]
    \item matrix multiplications in bf16 for speed and memory bandwidth;
    \item softmax, normalization reductions, and loss accumulation in float32;
    \item Adam moments and sometimes master weights in float32.
\end{itemize}

\subsection{Scaling-law perspective}

For a fixed compute budget, increasing model size without increasing training tokens can be inefficient. Compute-optimal scaling analyses suggest that model size and data size should be balanced rather than scaling only parameters \cite{hoffmann2022chinchilla}. For this project, if V1 improves validation loss over V0, the comparison should control for training tokens, optimizer settings, context length, and parameter count.

\subsection{Fair comparisons across Transformer variants}

To attribute performance differences to architecture rather than confounders, keep the following fixed whenever the mechanism permits it:
\begin{itemize}[nosep]
    \item tokenizer and vocabulary;
    \item training and validation data;
    \item number of training tokens;
    \item context length;
    \item optimizer and learning-rate schedule;
    \item token batch size;
    \item parameter count or active parameter count, as appropriate;
    \item evaluation frequency, precision policy, and hardware allocation.
\end{itemize}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=8mm and 7mm]
    \node[block] (v0) {V0 baseline};
    \node[block, right=of v0] (rms) {+ RMSNorm};
    \node[block, right=of rms] (rope) {+ RoPE};
    \node[block, below=of rope] (swiglu) {+ SwiGLU};
    \node[block, left=of swiglu] (flash) {+ FlashAttention};
    \node[data, left=of flash] (v1) {full V1};

    \draw[arrow] (v0) -- node[above, note] {scale} (rms);
    \draw[arrow] (rms) -- node[above, note] {position} (rope);
    \draw[arrow] (rope) -- node[right, note] {gating} (swiglu);
    \draw[arrow] (swiglu) -- node[below, note] {systems} (flash);
    \draw[arrow] (flash) -- (v1);
    \node[note, below=9mm of flash, align=center, text width=10cm] {Ablations isolate whether a gain comes from modeling quality, optimization stability, or computational efficiency.};
\end{tikzpicture}
\caption{Ablation path for a meaningful V0-to-V1 comparison. FlashAttention should mainly change speed and memory; RoPE, RMSNorm, and SwiGLU may affect loss.}
\label{fig:ablation-path}
\end{figure}



\subsection{Parameter parity is a control, not a complete notion of complexity}
\label{sec:param-parity}

Let $N_v$ be the trainable parameter count of variant $v$ and let
\begin{equation}
    \bar N=\frac{1}{|\mathcal{V}_{\mathrm{var}}|}
           \sum_{v\in\mathcal{V}_{\mathrm{var}}}N_v
\end{equation}
be the mean across the set of variants $\mathcal{V}_{\mathrm{var}}$. A parity rule such as
\begin{equation}\label{eq:param-parity}
    \frac{|N_v-\bar N|}{\bar N}\leq0.05
\end{equation}
limits raw-capacity differences.

This is useful because a much larger model may win simply through more parameters. It is not sufficient because parameter count does not determine computation or expressivity by itself. GQA removes K/V parameters but leaves much attention arithmetic unchanged. SWA changes pairwise attention FLOPs with no parameter change. MoE has distinct total and active parameter counts. Linear attention has the same projection parameters as MHA but a different state bottleneck. Fair reporting therefore includes parameters, active parameters, FLOPs, memory, and wall-clock time rather than treating any one as ``model complexity.''

\subsection{Three comparison axes: data, wall-clock time, and FLOPs}
\label{sec:three-axes}

One final validation loss answers only one question. The project uses three axes to separate statistical efficiency from systems efficiency.

Let a training log contain checkpoints
\begin{equation}
    \{(u_s,t_s,F_s,\widehat\loss_s)\}_{s=1}^{S},
\end{equation}
where $u_s$ is cumulative training tokens, $t_s$ is elapsed time, $F_s$ is cumulative estimated FLOPs, and $\widehat\loss_s$ is validation loss.

\begin{align}
    \widehat\loss^{\mathrm{data}}(U)
      &=\mathrm{Interp}\{(u_s,\widehat\loss_s)\}(U),\\
    \widehat\loss^{\mathrm{time}}(\tau)
      &=\mathrm{Interp}\{(t_s,\widehat\loss_s)\}(\tau),\\
    \widehat\loss^{\mathrm{FLOP}}(G)
      &=\mathrm{Interp}\{(F_s,\widehat\loss_s)\}(G).
\end{align}
Here $U$, $\tau$, and $G$ are common budgets, and $\mathrm{Interp}$ denotes an explicitly reported interpolation or smoothing rule.

\begin{center}
\begin{tabular}{lll}
\toprule
Axis & Held fixed & Main question\\
\midrule
Fixed data & tokens $U$ & which architecture learns most per token?\\
Fixed time & elapsed seconds $\tau$ & which gives the best result by a deadline?\\
Fixed FLOPs & compute budget $G$ & which extracts most learning per arithmetic operation?\\
\bottomrule
\end{tabular}
\end{center}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=9mm and 12mm]
    \node[block] (run) {same training logs};
    \node[block, above right=4mm and 15mm of run] (data) {slice at common\\token budget};
    \node[block, right=15mm of run] (time) {slice at common\\time budget};
    \node[block, below right=4mm and 15mm of run] (flop) {slice at common\\FLOP budget};
    \node[tensor, right=22mm of time] (compare) {three answers\\to three questions};
    \draw[arrow] (run)--(data); \draw[arrow] (run)--(time); \draw[arrow] (run)--(flop);
    \draw[arrow] (data)--(compare); \draw[arrow] (time)--(compare); \draw[arrow] (flop)--(compare);
\end{tikzpicture}
\caption{The same runs can be compared under three budgets. No single axis should be presented as the universally correct ranking.}
\label{fig:three-axes}
\end{figure}

Interpolation can hide noise or sparse logging. The common budget should lie inside every run's observed range, and the report should show the underlying curves. A fixed-time comparison also includes compiler, kernel, and dataloader effects; that is a feature when measuring practical performance, but a confound when making a purely mathematical architecture claim.

\subsection{Component-level FLOP accounting}
\label{sec:flop-accounting}

The approximation $6N\times\text{tokens}$ is useful for dense Transformers but cannot distinguish full attention from SWA, GQA projection savings, or sparse expert activation. A component model makes the assumptions visible.

For one forward pass with batch size $B$, length $T$, and a standard MHA layer, matrix-multiplication FLOPs are approximately
\begin{align}
    F_{QKV}&=6BT\dmodel^2,\\
    F_O&=2BT\dmodel^2,\\
    F_{\mathrm{SwiGLU}}&=6BT\dmodel d_h.
\end{align}
The factor $2$ in a matrix product counts one multiply and one add. SwiGLU has three projections: gate, up, and down.

Attention pairwise work is most clearly expressed using $P$, the number of allowed query--key pairs per sequence:
\begin{equation}\label{eq:attention-flops-pairs}
    F_{\mathrm{pair}}
    \approx4B\nhead\dhead P.
\end{equation}
The first factor $2$ is for $QK^\top$ and the second factor $2$ is for multiplying probabilities by $V$. For full causal attention,
\begin{equation}
    P=P_{\mathrm{full}}=\frac{T(T+1)}{2},
\end{equation}
and for SWA use \cref{eq:swa-pairs}. Actual kernel FLOPs may differ because tiles, padding, and masked regions can still incur work.

For GQA, replace the QKV term by
\begin{equation}
    F_{QKV}^{\mathrm{GQA}}
    =2BT\dmodel(\dmodel+2n_{\mathrm{kv}}\dhead),
\end{equation}
while pairwise attention still has $\nhead$ query heads. For MoE, the FFN term uses $k$ active experts per token, plus router cost; total expert parameters do not all contribute FLOPs to one token. For linear attention, the attention-state term is $O(BT\nhead r d_v)$ rather than a query--key pair count.

Backward propagation for a large matrix multiplication is roughly twice its forward cost, so a matmul-dominated training estimate often multiplies forward FLOPs by about three. This is an approximation, not a universal constant; nonlinearities, normalization, recomputation, optimizer updates, and kernel choices add work. Profiler measurements are the preferred validation.

\subsection{Model FLOPs utilization and the roofline model}
\label{sec:roofline}

Model FLOPs utilization (MFU) compares achieved model arithmetic with a hardware peak:
\begin{equation}\label{eq:mfu}
    \mathrm{MFU}
    =\frac{F_{\mathrm{model,step}}/t_{\mathrm{step}}}
           {P_{\mathrm{peak}}},
\end{equation}
where $F_{\mathrm{model,step}}$ is the chosen FLOP estimate, $t_{\mathrm{step}}$ is measured step time, and $P_{\mathrm{peak}}$ is a compatible peak throughput. Numerator and denominator must use the same precision and sparsity assumptions.

The roofline model relates performance to operational intensity \cite{williams2009roofline}:
\begin{equation}\label{eq:roofline}
    P_{\mathrm{bound}}(I)
    =\min(P_{\mathrm{peak}},\;\mathrm{BW}\cdot I),
    \qquad
    I=\frac{\text{FLOPs}}{\text{bytes transferred from DRAM}}.
\end{equation}
Here $\mathrm{BW}$ is sustained memory bandwidth, not merely a nominal interface speed. The ridge point is
\begin{equation}
    I^*=\frac{P_{\mathrm{peak}}}{\mathrm{BW}}.
\end{equation}
Below $I^*$, the bandwidth ceiling is lower; above $I^*$, compute peak is lower.

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=1.15cm,y=0.9cm]
    \draw[->, thick] (0,0)--(7.3,0) node[right] {operational intensity $I$};
    \draw[->, thick] (0,0)--(0,5.0) node[above] {performance $P$};
    \draw[very thick, blue!70!black] (0.3,0.35)--(4.2,4.15)--(7.0,4.15);
    \draw[dashed] (4.2,0)--(4.2,4.15);
    \node[note] at (2.1,2.5) {bandwidth bound\\$P=\mathrm{BW}\,I$};
    \node[note] at (5.7,4.5) {compute ceiling $P_{\mathrm{peak}}$};
    \node[note] at (4.2,-0.4) {ridge $I^*$};
    \fill (2.2,1.6) circle (2pt) node[below right, note] {memory-heavy variant};
    \fill (5.7,3.4) circle (2pt) node[below, note] {compute-heavy variant};
\end{tikzpicture}
\caption{Roofline model. It supplies an upper bound, not a prediction that every implementation reaches the roof.}
\label{fig:roofline}
\end{figure}

For the NVIDIA L4, the published BF16 tensor-core figure is 242 TFLOP/s \emph{with structured sparsity}; NVIDIA states that the non-sparse figure is half, about 121 TFLOP/s. Published memory bandwidth is 300 GB/s \cite{nvidial4spec}. Therefore the nominal dense ridge point is about
\begin{equation}
    I^*_{\mathrm{dense}}
    \approx\frac{121\times10^{12}}{300\times10^9}
    \approx403\ \text{FLOPs/byte},
\end{equation}
whereas 807 FLOPs/byte applies only if the workload genuinely uses the advertised structured-sparsity peak. A virtualized L4 profile may have additional scheduling limits, so a measured microbenchmark ceiling is preferable.

A point far below the roof can also be limited by launch overhead, insufficient parallelism, synchronization, or poor tensor shapes. Roofline classification is therefore a first diagnostic, not a complete performance explanation.

\subsection{Statistical reporting across random seeds}
\label{sec:seed-reporting}

Let $m_s$ be a metric from seed $s=1,\ldots,S$. Report the sample mean and standard deviation,
\begin{align}
    \bar m&=\frac{1}{S}\sum_{s=1}^{S}m_s,\\
    s_m&=\sqrt{\frac{1}{S-1}
        \sum_{s=1}^{S}(m_s-\bar m)^2}.
\end{align}
The standard deviation describes run-to-run variability. It is not the uncertainty of the mean. The standard error is
\begin{equation}
    \mathrm{SE}(\bar m)=\frac{s_m}{\sqrt S},
\end{equation}
and an approximate two-sided $(1-\gamma)$ confidence interval is
\begin{equation}
    \bar m\pm t_{1-\gamma/2,S-1}\frac{s_m}{\sqrt S},
\end{equation}
where $t_{q,\nu}$ is a Student-$t$ quantile with $\nu=S-1$ degrees of freedom.

With only three seeds, both $s_m$ and the confidence interval are unstable. The document should therefore show individual seed points and avoid treating overlapping or non-overlapping ``mean $\pm$ SD'' bars as a formal significance test. When variants can share data-order seeds, paired differences
\begin{equation}
    d_s=m_{A,s}-m_{B,s}
\end{equation}
usually reduce variance and answer the architectural comparison more directly.

\subsection{Stable rank: a scale-robust effective-dimension diagnostic}
\label{sec:stable-rank}

Collect $n$ token representations from layer $\ell$ into $\mH^{(\ell)}\in\reals^{n\times\dmodel}$. First center the columns:
\begin{equation}
    \widetilde{\mH}=\mC\mH,
    \qquad
    \mC=\mI_n-\frac{1}{n}\mathbf{1}\mathbf{1}^\top.
\end{equation}
For nonzero $\widetilde{\mH}$, stable rank is
\begin{equation}\label{eq:stable-rank}
    \mathrm{srank}(\widetilde{\mH})
    =\frac{\|\widetilde{\mH}\|_F^2}
           {\|\widetilde{\mH}\|_2^2}
    =\frac{\sum_j\sigma_j^2}{\sigma_1^2}.
\end{equation}
Here $\|\cdot\|_F$ is the Frobenius norm, $\|\cdot\|_2=\sigma_1$ is the spectral norm, and $\sigma_j$ are singular values. Stable rank lies between $1$ and the ordinary matrix rank.

If one singular direction dominates, the numerator and denominator are nearly equal and stable rank is near one. If energy is spread across $r$ equal singular directions, stable rank is $r$.

\begin{example}[Same ordinary rank, different effective rank]
Matrices with singular values $(10,1,1,1)$ and $(5,5,5,5)$ both have ordinary rank four. Their stable ranks are $(100+3)/100=1.03$ and $100/25=4$. Stable rank distinguishes a nearly one-dimensional representation from one that uses four directions comparably.
\end{example}

A falling curve can indicate concentration into a few directions, but high stable rank is not automatically better. A task may legitimately compress information into a low-dimensional sufficient statistic. Compare variants on the same data, centering rule, token sample, and layer location, and interpret stable rank alongside loss and retrieval probes.

\subsection{Centered kernel alignment: comparing representation geometry}
\label{sec:cka}

Suppose $\mX\in\reals^{n\times p}$ and $\mY\in\reals^{n\times q}$ contain representations of the same $n$ token examples from two layers or models. Center them using $\mC$:
\begin{equation}
    \mX_c=\mC\mX,
    \qquad
    \mY_c=\mC\mY.
\end{equation}
Linear centered kernel alignment (CKA) is
\begin{equation}\label{eq:cka}
    \mathrm{CKA}(\mX,\mY)
    =\frac{\|\mY_c^\top\mX_c\|_F^2}
           {\|\mX_c^\top\mX_c\|_F
            \|\mY_c^\top\mY_c\|_F}.
\end{equation}
It compares the pairwise geometry of examples and is invariant to isotropic scaling and orthogonal changes of feature basis \cite{kornblith2019cka}.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.72]
    \foreach \i in {0,...,5}{
      \foreach \j in {0,...,5}{
        \pgfmathsetmacro{\shade}{10+70*exp(-abs(\i-\j))}
        \fill[blue!\shade] (\j,5-\i) rectangle +(1,1);
      }
    }
    \draw[step=1cm, white, line width=0.4pt] (0,0) grid (6,6);
    \draw[thick] (0,0) rectangle (6,6);
    \node[below] at (3,-0.35) {layer in model B};
    \node[rotate=90] at (-0.45,3) {layer in model A};
    \node[note] at (3,6.35) {bright diagonal: corresponding depths are similar};
\end{tikzpicture}
\caption{Schematic CKA heatmap. The actual values are computed from centered hidden-state matrices; color here illustrates the interpretation.}
\label{fig:cka-heatmap}
\end{figure}

High adjacent-layer CKA does not by itself prove redundant computation. Residual connections make adjacent states share a large identity component, and a small but task-critical subspace may change while CKA remains high. More convincing evidence combines CKA with ablation, residual-update norms, and task metrics.

\subsection{Multi-query associative recall: a controlled retrieval probe}
\label{sec:mqar}

Multi-query associative recall (MQAR) tests whether a model can bind several key--value pairs in context and retrieve the correct value when each key reappears later \cite{arora2024zoology}. A synthetic sequence may contain
\begin{equation}
    (a_1,b_1),\ldots,(a_K,b_K)
\end{equation}
interleaved with distractors, followed by queries $a_{q_1},\ldots,a_{q_R}$. At a query for key $a_j$, the target next token is $b_j$.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.78, transform shape, node distance=3mm and 3mm]
    \node[data] (k1) {$a_1$}; \node[data, right=of k1] (v1) {$b_1$};
    \node[data, right=of v1] (d1) {noise};
    \node[data, right=of d1] (k2) {$a_2$}; \node[data, right=of k2] (v2) {$b_2$};
    \node[data, right=of v2] (d2) {$\cdots$};
    \node[data, right=of d2, fill=orange!12] (q2) {query $a_2$};
    \node[tensor, right=of q2] (pred) {predict $b_2$};
    \draw[arrow] (k1)--(v1); \draw[arrow] (v1)--(d1); \draw[arrow] (d1)--(k2); \draw[arrow] (k2)--(v2); \draw[arrow] (v2)--(d2); \draw[arrow] (d2)--(q2); \draw[arrow] (q2)--(pred);
    \draw[residual] (v2.north) .. controls +(0,1.1) and +(0,1.1) .. node[above, note] {association must survive distance and distractors} (q2.north);
\end{tikzpicture}
\caption{MQAR separates key--value binding and later retrieval, allowing distance and number of associations to be controlled.}
\label{fig:mqar}
\end{figure}

For query instance $r$, let $d_r$ be the token distance from its binding and let $y_r$ be the correct value. Accuracy in a distance bin $D$ is
\begin{equation}
    \mathrm{Acc}(D)
    =\frac{1}{|\{r:d_r\in D\}|}
      \sum_{r:d_r\in D}
      \ind\!\left[
        \argmax_{w} p_{\vtheta}(w\mid\text{context}_r)=y_r
      \right].
\end{equation}
Distance bins are usually more stable than requiring many examples at every exact integer distance. Report chance accuracy, vocabulary construction, number of associations, distractor process, and whether keys or values repeat.

MQAR isolates retrieval better than overall perplexity. SWA may degrade once the binding lies beyond direct or effective reach; fixed-state linear attention may degrade as more associations compete for the same state; full attention may remain stronger. It is still synthetic, so good MQAR performance is evidence of a mechanism, not proof of general language quality.

\subsection{Per-position context-use curves and a power-law summary}
\label{sec:icl-decay}

Let $L(t)$ be average validation NLL at sequence position $t$, computed over examples for which position $t$ is non-padding. Later positions have more preceding context, so $L(t)$ often decreases, but the curve can contain tokenization, document-boundary, and data-composition effects.

The project fits the descriptive model
\begin{equation}\label{eq:icl-power}
    L(t)=A t^{-\alpha}+C,
\end{equation}
where $A\geq0$ is the reducible amplitude, $\alpha\geq0$ controls the rate of decay, and $C\geq0$ is an asymptotic floor. A larger $\alpha$ means the fitted reducible part falls faster with position. It does \emph{not} automatically mean better long-context use: a curve can fall quickly and then hit a high floor $C$.

\begin{warningbox}[Treat the power law as a hypothesis, not a law]
Per-position loss is not guaranteed to follow \cref{eq:icl-power}. Report residual plots, confidence intervals, and $R^2$; exclude or separately model early boundary positions if they dominate the fit. Compare the full $L(t)$ curves in addition to the single exponent.
\end{warningbox}

A useful derived quantity is the fitted context benefit between positions $t_1<t_2$:
\begin{equation}
    \Delta L(t_1,t_2)
    =A(t_1^{-\alpha}-t_2^{-\alpha}).
\end{equation}
This uses all three parameters and is easier to interpret than $\alpha$ alone.

\subsection{Pareto fronts: reporting trade-offs without a single arbitrary score}
\label{sec:pareto}

Suppose each variant has a cost $x_v$ and a loss $y_v$, both to be minimized. Variant $v'$ dominates $v$ if
\begin{equation}
    x_{v'}\leq x_v,
    \qquad y_{v'}\leq y_v,
\end{equation}
with at least one strict inequality. The Pareto front is the set of non-dominated variants.

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=1.0cm,y=1.0cm]
    \draw[->, thick] (0,0)--(7,0) node[right] {cost};
    \draw[->, thick] (0,0)--(0,5.6) node[above] {validation loss};
    \foreach \x/\y/\lab in {1.0/4.7/A,1.8/3.8/B,2.8/3.2/C,4.2/2.7/D,5.6/2.5/E,3.8/4.1/F,5.1/3.5/G}{
      \fill (\x,\y) circle (2.5pt) node[above right, note] {\lab};
    }
    \draw[very thick, blue!70!black] (1.0,4.7)--(1.8,3.8)--(2.8,3.2)--(4.2,2.7)--(5.6,2.5);
    \node[note, blue!70!black] at (4.7,1.8) {non-dominated frontier};
    \draw[dashed, red!70!black, ->] (3.8,4.1)--(2.8,3.2);
    \node[note, red!70!black] at (4.8,4.35) {F is dominated by C};
\end{tikzpicture}
\caption{Pareto analysis. A point behind the frontier is worse on both objectives than at least one alternative.}
\label{fig:pareto}
\end{figure}

Useful objective pairs are cumulative FLOPs versus loss, wall-clock time versus loss, and peak memory versus loss. A variant can be Pareto-optimal without being the preferred choice; the user must still decide how much extra cost is worth a loss reduction. Seed uncertainty can also change dominance, so error bars or bootstrap dominance probabilities should accompany close points.


\subsection{Validation uncertainty, sanity checks, and calibration}

Token losses are correlated within documents, so a simple token-level standard error is too optimistic. A rougher batch- or document-level estimate is
\begin{equation}
    \mathrm{SE}(\widehat R)\approx \frac{s_{\mathrm{batch}}}{\sqrt{n_{\mathrm{batch}}}},
\end{equation}
where $s_{\mathrm{batch}}$ is the sample standard deviation of batch-level losses.

Before trusting a training run, verify:
\begin{enumerate}[nosep]
    \item loss at initialization is near $\log\vocab$ when logits are close to uniform;
    \item the model can overfit a tiny batch;
    \item predictions at position $t$ do not change when future tokens are modified;
    \item RoPE cosine/sine tensors broadcast correctly over batch and heads;
    \item cached generation logits match non-cached logits up to numerical tolerance;
    \item dropout, if used, is disabled during evaluation.
\end{enumerate}

Cross-entropy measures probabilistic accuracy, but a model can still be miscalibrated. Calibration can be checked by binning predictions by confidence and comparing average confidence with empirical accuracy.

% =============================================================================
\section{Computational Concepts Around the Model}
\label{sec:infrastructure}

\subsection{Graph compilation and kernel fusion}

A mathematical expression such as RMSNorm or SwiGLU is often implemented as several primitive operations. Launching each primitive separately incurs overhead and repeatedly moves intermediate tensors through memory. Graph compilation analyzes a region of computation and may fuse compatible operations so intermediates remain in registers or on-chip memory.

\begin{example}[Why fusion changes time but not the model]
The expression
\begin{equation*}
    \mW_{\mathrm{down}}\!\left[\SiLU(\mW_g\vx)\odot(\mW_u\vx)\right]
\end{equation*}
has the same mathematical value whether the activation and element-wise product are separate GPU kernels or one fused kernel. The statistical model is unchanged; only memory traffic, launch overhead, and numerical operation order may differ.
\end{example}

Static tensor shapes make optimization easier because the compiler can specialize schedules and memory plans. Dynamic generation shapes and growing caches can require multiple compiled graphs or reduce fusion opportunities. These effects belong in fixed-wall-clock and MFU analysis, not in claims about validation quality at a fixed token budget.

\subsection{Training, prefill, and decode throughput}

Training throughput is commonly
\begin{equation}
    \mathrm{tokens/sec}
    =\frac{B\,T\,S}{t_{\mathrm{elapsed}}},
\end{equation}
where $S$ is the number of optimizer steps. Report whether this counts target tokens, input tokens, gradient-accumulation micro-batches, and data-loading time.

Generation has two different regimes:
\begin{itemize}[nosep]
    \item \textbf{prefill:} processes the prompt in parallel and creates the initial cache;
    \item \textbf{decode:} produces one new token per sequence per iteration and repeatedly reads the cache.
\end{itemize}
Prefill resembles a training forward pass and can use large matrix products. Decode often becomes memory-bandwidth and latency sensitive because each step has little parallel work per sequence. GQA mainly helps decode cache traffic; FlashAttention mainly helps long parallel attention; these benefits should not be collapsed into one throughput number.

% =============================================================================
\section{How to Add Future Transformer Variants}
\label{sec:future-variants}
% =============================================================================

The project will later include more Transformer variants. New components should be added to the one section where they naturally belong, rather than creating a second summary section that repeats the explanation.

\subsection{Design-log template}

For each new component, add one subsection answering:
\begin{enumerate}[nosep]
    \item \textbf{Baseline role:} what did the old component do inside the model?
    \item \textbf{Limitation:} what became inefficient, unstable, or statistically weak in the project setting?
    \item \textbf{Design pressure:} was the constraint loss, memory, speed, context length, serving latency, or parameter efficiency?
    \item \textbf{Replacement mechanism:} what changes mathematically or computationally?
    \item \textbf{New trade-off:} what cost appears in parameters, memory, stability, interpretability, or implementation complexity?
    \item \textbf{Required re-tests:} which existing sanity checks or ablations must be repeated?
\end{enumerate}

\subsection{Where future components should go}

\begin{itemize}[nosep]
    \item Add scaled RoPE, position interpolation, or learned relative bias refinements to \cref{sec:attention-position}.
    \item Add dilated, block-sparse, retrieval-augmented, or state-space attention alternatives to \cref{sec:attention-position} or \cref{sec:generation}.
    \item Add ReGLU, GEGLU, expert-choice routing, or shared-expert variants to \cref{sec:ffn}.
    \item Add alternative normalizers or residual-scaling rules to \cref{sec:normalization}.
    \item Add fine-tuning regularizers, adapters, or bias changes to \cref{sec:simplifications} or \cref{sec:training-eval}.
\end{itemize}

\begin{motivationbox}[Rule for future notes]
Do not write ``we use component X because it is modern.'' Write ``we use component X because component Y creates limitation Z under constraint C, and X addresses Z by mechanism M while introducing trade-off T.''
\end{motivationbox}

% =============================================================================
\section{Final Evaluation and Static Report Corrections}
\label{sec:final-evaluation-corrections}
% =============================================================================

The completed main-scale runs exposed several reporting gaps that did not require
new training, but did require a stricter evaluator.  The final evaluator now treats
the three seeds as the unit of statistical comparison.  Fixed-data,
fixed-wall-clock, and fixed-FLOP losses are first computed at one shared global
budget and then reported as the mean and sample standard deviation across seeds.
The Pareto analysis uses those seed means rather than an arbitrary representative
checkpoint.  Diagnostic probes retain every checkpoint result in the raw schema
and separately expose a variant-level aggregate.  This prevents the last seed
processed from silently replacing the earlier seeds.

\subsection{Corrected parameter accounting}

The original report estimator used the nominal configuration value
$d_{\mathrm{ff}}=4d$ for SwiGLU.  The implemented module instead uses
$\lceil(8d/3)/64\rceil 64$, matching a conventional two-matrix $4d$ FFN after
adding the third gated projection.  The old estimator therefore overstated every
dense modern recipe and did not count MoE experts or routers at all.

The corrected accounting mirrors the instantiated modules.  It also separates
\emph{active parameters per token} from \emph{total stored parameters}.  For an
MoE layer, active count includes the router and the $k=2$ selected experts, while
total count includes all eight experts.  The resulting main-scale counts are:

\begin{table}[H]
\centering
\small
\begin{tabular}{lrr}
\toprule
Recipe & Active parameters & Total parameters \\
\midrule
Vanilla, Modern, ALiBi, SWA, SWA-interleaved, Linear & 51,430,400 & 51,430,400 \\
GQA & 48,284,672 & 48,284,672 \\
MoE & 68,764,672 & 172,573,696 \\
MoE-interleaved and MoE-deep & 60,097,536 & 112,002,048 \\
\bottomrule
\end{tabular}
\caption{Corrected parameter accounting for the saved main-scale recipes.
The parity test uses the active column; the report always displays both.}
\label{tab:corrected-parameter-counts}
\end{table}

The MoE recipes therefore do not satisfy the declared $\pm5\%$ active-parameter
parity rule.  This is now recorded as an experimental limitation rather than
hidden by the erroneous estimator.  No compensating retrain was introduced:
changing expert width, top-$k$, depth placement, or training tokens would define
a new experiment rather than repair the already completed one.

\subsection{Seed-aware serving and paired-tail long-context evaluation}

Serving remains a representative-checkpoint capability benchmark: seed 42 uses a
64-token prompt, eight generated tokens, batch size one, one warm-up, and two timed
repetitions on the NVIDIA L4-24Q.  Cached timing is reported only when the model
returns a genuinely reusable prompt cache, and persistent memory counts the tensor
storages retained by that opaque cache.  Serving timing is therefore not presented
as a seed-aggregated claim.

Long-context quality now uses all three independently trained checkpoints and eight
fixed, non-overlapping validation windows per checkpoint.  Each window contains
4097 tokens.  For a context length $c\in\{1024,2048,4096\}$, the evaluator crops
the window from the left so that every $c$ scores the same final $N=256$ target
tokens.  If $z_{s,c,w,t}$ denotes the logits for seed $s$, context $c$, window $w$,
and one of those shared tail positions, the checkpoint estimate is
\[
 \widehat{L}_{s,c}
 = \frac{1}{8N}\sum_{w=1}^{8}\sum_{t=1}^{N}
   \operatorname{CE}(z_{s,c,w,t}, y_{w,t}).
\]
The report first averages windows within a checkpoint, then reports the mean and
sample standard deviation of the three checkpoint estimates.  Paired degradation
is computed within each seed before aggregation.  This makes the checkpoint seed,
not a token or window, the independent uncertainty unit.  Target-free prefill
timing ends before float32 cross-entropy is computed, so quality work is not
mislabelled as inference throughput.

\begin{table}[H]
\centering
\scriptsize
\begin{tabular}{lrrrrr}
\toprule
Recipe & Tail PPL@1K & Tail PPL@2K & Tail PPL@4K & 4K/1K & 4K ktok/s \\
\midrule
ALiBi & $56.07\pm0.51$ & $58.29\pm0.88$ & $58.74\pm1.10$ & $1.047\pm0.010$ & 242.6 \\
SWA & $58.95\pm0.45$ & $58.95\pm0.48$ & $58.93\pm0.46$ & $1.000\pm0.001$ & 253.4 \\
SWA-interleaved & $58.58\pm0.28$ & $78.48\pm4.50$ & $84.86\pm6.12$ & $1.448\pm0.100$ & 237.9 \\
GQA & $60.10\pm0.41$ & $370.72\pm82.01$ & $437.37\pm47.76$ & $7.279\pm0.815$ & 239.6 \\
MoE-deep & $34.10\pm0.65$ & $458.63\pm61.52$ & $488.85\pm23.26$ & $14.329\pm0.423$ & 42.0 \\
MoE-interleaved & $34.40\pm0.66$ & $463.84\pm54.23$ & $518.29\pm18.30$ & $15.077\pm0.817$ & 41.7 \\
MoE & $31.91\pm0.29$ & $427.35\pm24.78$ & $526.63\pm17.94$ & $16.505\pm0.508$ & 36.0 \\
Modern & $39.80\pm0.31$ & $453.01\pm61.98$ & $553.94\pm18.95$ & $13.917\pm0.368$ & 51.2 \\
Linear & $50.10\pm0.54$ & $1103.41\pm299.49$ & $1311.78\pm25.66$ & $26.186\pm0.789$ & 34.7 \\
Vanilla & $44.52\pm0.39$ & unsupported & unsupported & unsupported & unsupported \\
\bottomrule
\end{tabular}
\caption{Paired-tail long-context quality.  Each entry is mean $\pm$ sample
standard deviation over three checkpoint seeds; every checkpoint estimate averages
eight windows and 2048 shared target predictions.  Throughput is the cross-seed
mean of target-free 4096-token prefill.}
\label{tab:inference-long-context-final}
\end{table}

ALiBi has the lowest mean 4096-token tail perplexity, but SWA is only 0.20
perplexity points behind.  That gap is smaller than either model's cross-seed
standard deviation, so these data do not justify an absolute-quality winner.
SWA has the strongest retention ($0.9998\pm0.0008$) and the faster 4096-token
prefill of the two.  ALiBi degrades only $4.7\%$, while SWA-interleaved degrades
about $44.8\%$.  GQA, the dense/MoE RoPE recipes, and causal linear attention
all deteriorate sharply outside the training length; causal linear attention is
worst at a $26.19\times$ perplexity ratio.  Linear asymptotic complexity therefore
does not imply length generalization.

SWA's flat curve needs a careful interpretation.  Its receptive field remains
256 tokens, so stability at a 4096-token input shows that the trained local
computation transfers; it does not show that the model retrieves information
thousands of tokens away.  ALiBi and SWA are consequently co-leaders in absolute
tail quality, SWA is the stability leader, and long-range retrieval remains a
separate question.  Vanilla is explicitly unsupported beyond its learned
1024-position table.  ALiBi, GQA, both SWA recipes, and causal linear attention
still expose no reusable prompt cache through the shared interface; those serving
paths remain labelled unsupported rather than emulated.


\subsection{Seed provenance and fresh final validation}

The final evaluator no longer treats a copied training log as an independent
checkpoint result.  It detects byte-equivalent parsed \texttt{metrics.jsonl}
histories within a recipe and persists an explicit warning in report metadata,
which the HTML dashboard renders in its provenance section.  Historical wall-clock
and pre-endpoint fixed-FLOP curves from such logs retain their mean for reference
but expose no standard deviation.  Endpoint fixed-data, Pareto, and aggregate
quality values instead use a fresh 50-batch held-out validation pass over each
saved checkpoint;
their sample standard deviations therefore reflect the three independently
trained weight states.  MQAR, stable-rank, CKA, and attention-entropy JSON now
retain elementwise cross-seed sample standard deviations alongside their means.

The canonical V5 inputs for this report are
\texttt{linear\_main\_1B\_fixed\_s42},
\texttt{linear\_main\_1B\_fixed\_s137}, and
\texttt{linear\_main\_1B\_fixed\_s2024}.  The older unsuffixed V5
directories contain the superseded Linformer projection tensors
\texttt{E}/\texttt{F} and are deliberately excluded from final evaluation.

\subsection{Self-contained HTML dashboard}

The Streamlit application is no longer the supported report surface.  The
evaluation pipeline now writes a single \texttt{index.html}; the standalone
\texttt{scripts/build\_dashboard.py} command can rebuild it after the inference
benchmark.  Metrics, benchmark statuses, metadata, CSS, JavaScript, and PNG plots
are embedded directly in the file.  It opens without a Python server, network
connection, CDN, or tracking runtime.  The versioned JSON remains the machine
interface, while the HTML is a portable presentation artifact.


% =============================================================================
\section{Operational Hardening and Static-Site Assets}

\subsection{Fault-tolerant training path}

The fault-tolerance components described earlier are now connected to the public
training command.  Passing \texttt{--fault-tolerant} constructs a
\texttt{CheckpointRingBuffer}, \texttt{AsyncCheckpointWriter}, and
\texttt{HealthMonitor} and injects them into the trainer.  The operating sequence is
therefore:
\begin{enumerate}
    \item copy model and optimizer state to CPU-owned tensors;
    \item serialize in a single background worker to a temporary file;
    \item flush the file with \texttt{fsync}, atomically rename it, and compute SHA-256;
    \item register the checkpoint only after integrity verification; and
    \item evict the oldest ring member only after the replacement is valid.
\end{enumerate}
Training waits for the final queued save before returning, so command completion is a
durability boundary rather than merely a request to save.

The fault-tolerant schema stores model and optimizer state plus token count, best
validation loss, skipped-step count, exact training and validation loader cursors,
and Python, PyTorch, and CUDA random-number generator state.  Resume accepts either
an explicit verified path or
\texttt{--resume latest}.  The latter scans the persisted ring from newest to oldest
and compares each file against the hash in ring metadata; a bit-flipped newest file is
therefore skipped in favor of the newest valid predecessor.  This distinction matters:
fault-tolerant explicit resume requires a pre-existing SHA-256 sidecar instead of
trusting bytes without a previously persisted expected digest, and automatic ring
recovery checks the independent digest recorded in ring metadata.

The health monitor observes the un-clipped gradient norm and loss.  Non-finite values
abort the optimizer step before clipping can contaminate otherwise finite gradients.
Finite statistical spikes skip the step; repeated spikes escalate to a rollback.  A
rollback reloads model and optimizer state, progress counters, loader cursors, and RNG
state, then resets the monitor window.  Fault-injection tests exercise corruption,
interrupted
writes, NaN/Inf handling, spike escalation, rotation, asynchronous completion, and
crash-style state restoration.  Snapshot latency remains hardware dependent, so the
original 200 ms target is treated as an operational measurement to report on the
deployment GPU rather than a cross-platform guarantee.

\subsection{Portable model-internals export}

The publication frontend will live in a separate GitHub Pages/Jekyll repository.
Consequently, \texttt{scripts/export\_site\_assets.py} emits a frontend-agnostic
bundle instead of another application runtime.  Its manifest points to strict JSON
containing cross-seed stable-rank curves, adjacent and full CKA matrices, attention
entropy from the aggregate probe where available, captured-context attention entropy
for every supported softmax variant, and attention-pattern records by layer and head.
PNG curves and mean-head heatmaps provide static fallbacks.  A site can therefore
implement interactive
variant, layer, and head selectors without importing the training repository or
running Python.

Memory-efficient attention kernels deliberately avoid materializing the quadratic
probability matrix.  For visualization only, the exporter captures the normalized
input to each selected attention module and reconstructs its pre-dropout probabilities:
\[
    A_h = \operatorname{softmax}\!\left(
        \frac{Q_h K_h^\top}{\sqrt{d_h}} + B_h + M
    \right).
\]
Here \(B_h\) is the ALiBi bias when applicable and \(M\) is the causal mask,
augmented by the local-window constraint for SWA layers.  RoPE is applied before the
dot product, and GQA key heads are repeated across their associated query-head groups.
This is mathematically equivalent to the probabilities used by full softmax attention,
SDPA, and FlashAttention before dropout, even though the production kernel never
stores \(A_h\).  Tests verify that rows sum to one and that all mass above the causal
diagonal is zero.

\begin{warningbox}[Why V5 has no pairwise attention heatmap]
Causal linear attention computes outputs from prefix feature-map statistics rather
than a normalized pairwise softmax matrix.  Its rotated numerator and unrotated
denominator do not define a single conventional \(T\!\times\!T\) probability map.
The asset schema therefore records V5 as \texttt{unsupported} with this reason.  It
does not invent a softmax visualization that the trained model never used.
\end{warningbox}

\section{References}
% =============================================================================

\begin{thebibliography}{99}

\bibitem{vaswani2017attention}
Ashish Vaswani, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion Jones, Aidan N. Gomez, Lukasz Kaiser, and Illia Polosukhin.
\newblock Attention Is All You Need.
\newblock \emph{NeurIPS}, 2017.

\bibitem{ba2016layernorm}
Jimmy Lei Ba, Jamie Ryan Kiros, and Geoffrey E. Hinton.
\newblock Layer Normalization.
\newblock arXiv:1607.06450, 2016.

\bibitem{radford2019gpt2}
Alec Radford, Jeffrey Wu, Rewon Child, David Luan, Dario Amodei, and Ilya Sutskever.
\newblock Language Models are Unsupervised Multitask Learners.
\newblock OpenAI technical report, 2019.

\bibitem{zhang2019rmsnorm}
Biao Zhang and Rico Sennrich.
\newblock Root Mean Square Layer Normalization.
\newblock \emph{NeurIPS}, 2019.

\bibitem{su2021roformer}
Jianlin Su, Yu Lu, Shengfeng Pan, Ahmed Murtadha, Bo Wen, and Yunfeng Liu.
\newblock RoFormer: Enhanced Transformer with Rotary Position Embedding.
\newblock arXiv:2104.09864, 2021.

\bibitem{hendrycks2016gelu}
Dan Hendrycks and Kevin Gimpel.
\newblock Gaussian Error Linear Units (GELUs).
\newblock arXiv:1606.08415, 2016.

\bibitem{shazeer2020glu}
Noam Shazeer.
\newblock GLU Variants Improve Transformer.
\newblock arXiv:2002.05202, 2020.

\bibitem{dao2022flashattention}
Tri Dao, Daniel Y. Fu, Stefano Ermon, Atri Rudra, and Christopher Re.
\newblock FlashAttention: Fast and Memory-Efficient Exact Attention with IO-Awareness.
\newblock \emph{NeurIPS}, 2022.

\bibitem{touvron2023llama}
Hugo Touvron, Thibaut Lavril, Gautier Izacard, Xavier Martinet, Marie-Anne Lachaux, Timothee Lacroix, Baptiste Roziere, Naman Goyal, Eric Hambro, Faisal Azhar, Aurelien Rodriguez, Armand Joulin, Edouard Grave, and Guillaume Lample.
\newblock LLaMA: Open and Efficient Foundation Language Models.
\newblock arXiv:2302.13971, 2023.

\bibitem{loshchilov2019adamw}
Ilya Loshchilov and Frank Hutter.
\newblock Decoupled Weight Decay Regularization.
\newblock \emph{ICLR}, 2019.

\bibitem{press2017weighttying}
Ofir Press and Lior Wolf.
\newblock Using the Output Embedding to Improve Language Models.
\newblock \emph{EACL}, 2017.

\bibitem{fedus2022switch}
William Fedus, Barret Zoph, and Noam Shazeer.
\newblock Switch Transformers: Scaling to Trillion Parameter Models with Simple and Efficient Sparsity.
\newblock \emph{JMLR}, 2022.

\bibitem{jiang2024mixtral}
Albert Q. Jiang, Alexandre Sablayrolles, Antoine Roux, Arthur Mensch, Blanche Savary, Chris Bamford, Devendra Singh Chaplot, Diego de las Casas, et al.
\newblock Mixtral of Experts.
\newblock arXiv:2401.04088, 2024.

\bibitem{zoph2022stmoe}
Barret Zoph, Irwan Bello, Sameer Kumar, Nan Du, Yanping Huang, Jeff Dean, Noam Shazeer, and William Fedus.
\newblock ST-MoE: Designing Stable and Transferable Sparse Expert Models.
\newblock arXiv:2202.08906, 2022.

\bibitem{holtzman2019nucleus}
Ari Holtzman, Jan Buys, Li Du, Maxwell Forbes, and Yejin Choi.
\newblock The Curious Case of Neural Text Degeneration.
\newblock arXiv:1904.09751, 2019.

\bibitem{hoffmann2022chinchilla}
Jordan Hoffmann, Sebastian Borgeaud, Arthur Mensch, Elena Buchatskaya, Trevor Cai, Eliza Rutherford, Diego de Las Casas, Lisa Anne Hendricks, Johannes Welbl, Aidan Clark, Tom Hennigan, Eric Noland, Katie Millican, George van den Driessche, Bogdan Damoc, Aurelia Guy, Simon Osindero, Karen Simonyan, Erich Elsen, Jack W. Rae, Oriol Vinyals, and Laurent Sifre.
\newblock Training Compute-Optimal Large Language Models.
\newblock arXiv:2203.15556, 2022.

\bibitem{press2022alibi}
Ofir Press, Noah A. Smith, and Mike Lewis.
\newblock Train Short, Test Long: Attention with Linear Biases Enables Input Length Extrapolation.
\newblock \emph{ICLR}, 2022.

\bibitem{ainslie2023gqa}
Joshua Ainslie, James Lee-Thorp, Michiel de Jong, Yury Zemlyanskiy, Federico Lebr\'on, and Sumit Sanghai.
\newblock GQA: Training Generalized Multi-Query Transformer Models from Multi-Head Checkpoints.
\newblock \emph{EMNLP}, 2023.

\bibitem{shazeer2019mqa}
Noam Shazeer.
\newblock Fast Transformer Decoding: One Write-Head is All You Need.
\newblock arXiv:1911.02150, 2019.

\bibitem{dao2023flashattention2}
Tri Dao.
\newblock FlashAttention-2: Faster Attention with Better Parallelism and Work Partitioning.
\newblock \emph{ICLR}, 2024.

\bibitem{katharopoulos2020linear}
Angelos Katharopoulos, Apoorv Vyas, Nikolaos Pappas, and Fran\c{c}ois Fleuret.
\newblock Transformers are RNNs: Fast Autoregressive Transformers with Linear Attention.
\newblock \emph{ICML}, 2020.

\bibitem{qin2022devil}
Zhen Qin, Weixuan Sun, Hui Deng, Dongxu Li, Yunshen Wei, Baohong Lv, Junjie Yan, Lingpeng Kong, and Yiran Zhong.
\newblock The Devil in Linear Transformer.
\newblock \emph{EMNLP}, 2022.

\bibitem{jiang2023mistral}
Albert Q. Jiang, Alexandre Sablayrolles, Arthur Mensch, Chris Bamford, Devendra Singh Chaplot, Diego de Las Casas, Florian Bressand, Gianna Lengyel, Guillaume Lample, Lucile Saulnier, L\'{e}lio Renard Lavaud, Marie-Anne Lachaux, Pierre Stock, Teven Le Scao, Thibaut Lavril, Thomas Wang, Timoth\'{e}e Lacroix, and William El Sayed.
\newblock Mistral 7B.
\newblock arXiv:2310.06825, 2023.

\bibitem{beltagy2020longformer}
Iz Beltagy, Matthew E. Peters, and Arman Cohan.
\newblock Longformer: The Long-Document Transformer.
\newblock arXiv:2004.05150, 2020.



\bibitem{gemma2024}
Gemma Team et al.
\newblock Gemma 2: Improving Open Language Models at a Practical Size.
\newblock arXiv:2408.00118, 2024.

\bibitem{kornblith2019cka}
Simon Kornblith, Mohammad Norouzi, Honglak Lee, and Geoffrey Hinton.
\newblock Similarity of Neural Network Representations Revisited.
\newblock \emph{ICML}, 2019.

\bibitem{arora2024zoology}
Simran Arora, Sabri Eyuboglu, Aman Timalsina, Isys Johnson, Michael Poli, James Zou, Atri Rudra, and Christopher R\'e.
\newblock Zoology: Measuring and Improving Recall in Efficient Language Models.
\newblock arXiv:2312.04927, 2023.

\bibitem{williams2009roofline}
Samuel Williams, Andrew Waterman, and David Patterson.
\newblock Roofline: An Insightful Visual Performance Model for Multicore Architectures.
\newblock \emph{Communications of the ACM}, 52(4):65--76, 2009.

\bibitem{nvidial4spec}
NVIDIA.
\newblock NVIDIA L4 Tensor Core GPU Product Specifications.
\newblock \url{https://www.nvidia.com/en-us/data-center/l4/}, accessed July 2026.


\end{thebibliography}

% =============================================================================

\end{document}
